### 2025-02-13 paper.tex
Start CYOA paper, write intro

\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
\hideLIPIcs

\bibliographystyle{plainurl}
\title{The Choose-Your-Own-Adventure Calculus} % (Pearl)

\author{Tomas Petricek}{Charles University, Prague, Czechia}{tomas@tomasp.net}{0000-0002-7242-2208}{}
\author{Jan Liam Verter}{Charles University, Prague, Czechia}{todo@todo}{}{}
\author{Mikoláš Fromm}{Charles University, Prague, Czechia}{todo@todo}{}{}
\ccsdesc[500]{Software and its engineering~General programming languages}
\keywords{Interactive programming systems, Type providers, Proof assistants}
\nolinenumbers

% \funding{funding}
% \acknowledgements{I want to thank \dots}
\authorrunning{Authors}

\begin{document}

\maketitle

\begin{abstract}

Some of the most remarkable results in mathematics reveal surprising connections between different
branches of the discipline. The aim of this paper is to point out a modest, but still
remarkable, similarity between a range of different interactive programming systems.

~

use a simple formal mathematical model
that we call \emph{the choose-your-own-adventure calculus} to

todo

% and enable a transfer of ideas between them

\end{abstract}

\newpage

\section{Introduction}

Multiple interactive programming systems, ranging from code editors for object-oriented programming
languages to data exploration systems and interactive proof assistants, exhibit a remarkably
similar pattern of interaction. They offer the user, who can be a programmer, a data scientist
or a proof writer, a range of choices that the user can select from in order to complete their
program, script or proof. The user can initiate the interaction iteratively, using it to
create and refine a larger part of their program.

There are subtle differences between different implementations of the general pattern. In some
systems, the resulting source code will contain a trace of the choices made by the user.
For example, when choosing an item from a list of class members, the code will contain the member
name. In some systems, the interaction results in a block of code that can be included in the
source file, but does not include a trace of the interaction. For example, invoking a proof search
or case split in Idris \cite{brady-2015-idris} constructs a well-typed program, but leaves no trace
of the command used to construct it.
%
The nature of the generated options also varies. The list of choices may include all possible
options that are valid at a given location, or it may list only a subset of the valid options.
In some cases, it may also include incorrect options as, for example, in auto-completion
for dynamic languages \cite{frolich-2024-autocomplete}.

The aim of this paper is paper is to formally capture the recurring interaction pattern:

\begin{enumerate}
\setlength{\itemsep}{5pt}
\item We motivate the formalism by reviewing four different systems that implement a variation on the
interaction pattern. These include type providers for data access in F\# \cite{syme-2013-inforich},
type providers for data exploration in The Gamma \cite{petricek-2022-thegamma,petricek-2017-dotdriven},
AI assistants for semi-automated data wrangling \cite{petricek-2023-aias} and tooling for
interactive proof assistants \cite{altenkirch-1994-alf,brady-2015-idris,verter-2024-mixed} (Section~\ref{sec:motivation}).

\item We introduce the \emph{choose-your-own-adventure calculus}, which is a small formal structure
that models an interactive system where a user constructs a program by repeatedly choosing from a list
of options offered by the system (Section~\ref{sec:calculus}).

\item The calculus allows us to make the aforementioned subtle differences precise. We define the
notions of \emph{correctness} and \emph{completeness} for the choose-your-own-adventure calculus.
To distinguish the different ways of embedding the interactions in the edited programs, we
also formally define \emph{base-level} and \emph{meta-level} integration of the system.

\item We show that various programmer assistance tools, such as search and AI-based recommendations
can be built on top of the primitives offered by the calculus, showing how the
choose-your-own-adventure calculus supports of transfer of ideas across different kinds of
interactive programming systems.
\end{enumerate}

The main contribution of this paper is conceptual rather than technical. Programming language
researchers have long focused on \emph{programming languages} rather than on a more general notion
of interactive \emph{programming systems} \cite{rpg-2012-revolution,jakubovic-2023-techdims}.
However, programming language theory methods can be equally useful for studying interactive
systems~\cite{adams-2025-grove}. The choose-your-own-adventure-calculus is another step in this
direction. It brings to light a pattern that is perhaps not surprising, but is very easy to
overlook until it is given a name.

\newpage

\section{Motivation}
\label{sec:motivation}

~

\section{Formal model}
\label{sec:calculus}

~

\section{Examples}
\label{sec:examples}

~

\section{Properties}
\label{sec:properties}

~

\section{Applications}
\label{sec:applications}

\cite{jakubovic-2023-techdims}
\cite{petricek-2020-live}

\newpage

\bibliography{paper}

\newpage

\section{Typesetting instructions -- Summary}
\label{sec:typesetting-summary}

LIPIcs is a series of open access high-quality conference proceedings across all fields in informatics established in cooperation with Schloss Dagstuhl.
In order to do justice to the high scientific quality of the conferences that publish their proceedings in the LIPIcs series, which is ensured by the thorough review process of the respective events, we believe that LIPIcs proceedings must have an attractive and consistent layout matching the standard of the series.
Moreover, the quality of the metadata, the typesetting and the layout must also meet the requirements of other external parties such as indexing service, DOI registry, funding agencies, among others. The guidelines contained in this document serve as the baseline for the authors, editors, and the publisher to create documents that meet as many different requirements as possible.

Please comply with the following instructions when preparing your article for a LIPIcs proceedings volume.
\paragraph*{Minimum requirements}

\begin{itemize}
\item Use pdflatex and an up-to-date \LaTeX{} system.
\item Use further \LaTeX{} packages and custom made macros carefully and only if required.
\item Use the provided sectioning macros: \verb+\section+, \verb+\subsection+, \verb+\subsubsection+, \linebreak \verb+\paragraph+, \verb+\paragraph*+, and \verb+\subparagraph*+.
\item Provide suitable graphics of at least 300dpi (preferably in PDF format).
\item Use BibTeX and keep the standard style (\verb+plainurl+) for the bibliography.
\item Please try to keep the warnings log as small as possible. Avoid overfull \verb+\hboxes+ and any kind of warnings/errors with the referenced BibTeX entries.
\item Use a spellchecker to correct typos.
\end{itemize}

\paragraph*{Mandatory metadata macros}
Please set the values of the metadata macros carefully since the information parsed from these macros will be passed to publication servers, catalogues and search engines.
Avoid placing macros inside the metadata macros. The following metadata macros/environments are mandatory:
\begin{itemize}
\item \verb+\title+ and, in case of long titles, \verb+\titlerunning+.
\item \verb+\author+, one for each author, even if two or more authors have the same affiliation.
\item \verb+\authorrunning+ and \verb+\Copyright+ (concatenated author names)\\
The \verb+\author+ macros and the \verb+\Copyright+ macro should contain full author names (especially with regard to the first name), while \verb+\authorrunning+ should contain abbreviated first names.
\item \verb+\ccsdesc+ (ACM classification, see \url{https://www.acm.org/publications/class-2012}).
\item \verb+\keywords+ (a comma-separated list of keywords).
\item \verb+\relatedversion+ (if there is a related version, typically the ``full version''); please make sure to provide a persistent URL, e.\,g., at arXiv.
\item \verb+\begin{abstract}...\end{abstract}+ .
\end{itemize}

\paragraph*{Please do not \ldots} %Do not override the \texttt{\seriesstyle}-defaults}
Generally speaking, please do not override the \texttt{lipics-v2021}-style defaults. To be more specific, a short checklist also used by Dagstuhl Publishing during the final typesetting is given below.
In case of \textbf{non-compliance} with these rules Dagstuhl Publishing will remove the corresponding parts of \LaTeX{} code and \textbf{replace it with the \texttt{lipics-v2021} defaults}. In serious cases, we may reject the LaTeX-source and expect the corresponding author to revise the relevant parts.
\begin{itemize}
\item Do not use a different main font. (For example, the \texttt{times} package is forbidden.)
\item Do not alter the spacing of the \texttt{lipics-v2021.cls} style file.
\item Do not use \verb+enumitem+ and \verb+paralist+. (The \texttt{enumerate} package is preloaded, so you can use
 \verb+\begin{enumerate}[(a)]+ or the like.)
\item Do not use ``self-made'' sectioning commands (e.\,g., \verb+\noindent{\bf My+ \verb+Paragraph}+).
\item Do not hide large text blocks using comments or \verb+\iffalse+ $\ldots$ \verb+\fi+ constructions.
\item Do not use conditional structures to include/exclude content. Instead, please provide only the content that should be published -- in one file -- and nothing else.
\item Do not wrap figures and tables with text. In particular, the package \texttt{wrapfig} is not supported.
\item Do not change the bibliography style. In particular, do not use author-year citations. (The
\texttt{natbib} package is not supported.)
\end{itemize}

\enlargethispage{\baselineskip}

This is only a summary containing the most relevant details. Please read the complete document ``LIPIcs: Instructions for Authors and the \texttt{lipics-v2021} Class'' for all details and don't hesitate to contact Dagstuhl Publishing (\url{mailto:publishing@dagstuhl.de}) in case of questions or comments:
\href{http://drops.dagstuhl.de/styles/lipics-v2021/lipics-v2021-authors/lipics-v2021-authors-guidelines.pdf}{\texttt{http://drops.dagstuhl.de/styles/lipics-v2021/\newline lipics-v2021-authors/lipics-v2021-authors-guidelines.pdf}}

\section{Lorem ipsum dolor sit amet}

Lorem ipsum dolor sit amet, consectetur adipiscing elit \cite{DBLP:journals/cacm/Knuth74}. Praesent convallis orci arcu, eu mollis dolor. Aliquam eleifend suscipit lacinia. Maecenas quam mi, porta ut lacinia sed, convallis ac dui. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Suspendisse potenti. Donec eget odio et magna ullamcorper vehicula ut vitae libero. Maecenas lectus nulla, auctor nec varius ac, ultricies et turpis. Pellentesque id ante erat. In hac habitasse platea dictumst. Curabitur a scelerisque odio. Pellentesque elit risus, posuere quis elementum at, pellentesque ut diam. Quisque aliquam libero id mi imperdiet quis convallis turpis eleifend.

\begin{lemma}[Lorem ipsum]
\label{lemma:lorem}
Vestibulum sodales dolor et dui cursus iaculis. Nullam ullamcorper purus vel turpis lobortis eu tempus lorem semper. Proin facilisis gravida rutrum. Etiam sed sollicitudin lorem. Proin pellentesque risus at elit hendrerit pharetra. Integer at turpis varius libero rhoncus fermentum vitae vitae metus.
\end{lemma}

\begin{proof}
Cras purus lorem, pulvinar et fermentum sagittis, suscipit quis magna.

\proofsubparagraph*{Just some paragraph within the proof.}
Nam liber tempor cum soluta nobis eleifend option congue nihil imperdiet doming id quod mazim placerat facer possim assum. Lorem ipsum dolor sit amet, consectetuer adipiscing elit, sed diam nonummy nibh euismod tincidunt ut laoreet dolore magna aliquam erat volutpat.
\begin{claim}
content...
\end{claim}
\begin{claimproof}
content...
    \begin{enumerate}
        \item abc abc abc \claimqedhere{}
    \end{enumerate}
\end{claimproof}

\end{proof}

\begin{corollary}[Curabitur pulvinar, \cite{DBLP:books/mk/GrayR93}]
\label{lemma:curabitur}
Nam liber tempor cum soluta nobis eleifend option congue nihil imperdiet doming id quod mazim placerat facer possim assum. Lorem ipsum dolor sit amet, consectetuer adipiscing elit, sed diam nonummy nibh euismod tincidunt ut laoreet dolore magna aliquam erat volutpat.
\end{corollary}

\begin{proposition}\label{prop1}
This is a proposition
\end{proposition}

\autoref{prop1} and \cref{prop1} \ldots

\subsection{Curabitur dictum felis id sapien}

Curabitur dictum \cref{lemma:curabitur} felis id sapien \autoref{lemma:curabitur} mollis ut venenatis tortor feugiat. Curabitur sed velit diam. Integer aliquam, nunc ac egestas lacinia, nibh est vehicula nibh, ac auctor velit tellus non arcu. Vestibulum lacinia ipsum vitae nisi ultrices eget gravida turpis laoreet. Duis rutrum dapibus ornare. Nulla vehicula vulputate iaculis. Proin a consequat neque. Donec ut rutrum urna. Morbi scelerisque turpis sed elit sagittis eu scelerisque quam condimentum. Pellentesque habitant morbi tristique senectus et netus et malesuada fames ac turpis egestas. Aenean nec faucibus leo. Cras ut nisl odio, non tincidunt lorem. Integer purus ligula, venenatis et convallis lacinia, scelerisque at erat. Fusce risus libero, convallis at fermentum in, dignissim sed sem. Ut dapibus orci vitae nisl viverra nec adipiscing tortor condimentum \cite{DBLP:journals/cacm/Dijkstra68a}. Donec non suscipit lorem. Nam sit amet enim vitae nisl accumsan pretium.

\begin{lstlisting}[caption={Useless code.},label=list:8-6,captionpos=t,float,abovecaptionskip=-\medskipamount]
for i:=maxint to 0 do
begin
    j:=square(root(i));
end;
\end{lstlisting}

\subsection{Proin ac fermentum augue}

Proin ac fermentum augue. Nullam bibendum enim sollicitudin tellus egestas lacinia euismod orci mollis. Nulla facilisi. Vivamus volutpat venenatis sapien, vitae feugiat arcu fringilla ac. Mauris sapien tortor, sagittis eget auctor at, vulputate pharetra magna. Sed congue, dui nec vulputate convallis, sem nunc adipiscing dui, vel venenatis mauris sem in dui. Praesent a pretium quam. Mauris non mauris sit amet eros rutrum aliquam id ut sapien. Nulla aliquet fringilla sagittis. Pellentesque eu metus posuere nunc tincidunt dignissim in tempor dolor. Nulla cursus aliquet enim. Cras sapien risus, accumsan eu cursus ut, commodo vel velit. Praesent aliquet consectetur ligula, vitae iaculis ligula interdum vel. Integer faucibus faucibus felis.

\begin{itemize}
\item Ut vitae diam augue.
\item Integer lacus ante, pellentesque sed sollicitudin et, pulvinar adipiscing sem.
\item Maecenas facilisis, leo quis tincidunt egestas, magna ipsum condimentum orci, vitae facilisis nibh turpis et elit.
\end{itemize}

\begin{remark}
content...
\end{remark}

\section{Pellentesque quis tortor}

Nec urna malesuada sollicitudin. Nulla facilisi. Vivamus aliquam tempus ligula eget ornare. Praesent eget magna ut turpis mattis cursus. Aliquam vel condimentum orci. Nunc congue, libero in gravida convallis \cite{DBLP:conf/focs/HopcroftPV75}, orci nibh sodales quam, id egestas felis mi nec nisi. Suspendisse tincidunt, est ac vestibulum posuere, justo odio bibendum urna, rutrum bibendum dolor sem nec tellus.

\begin{lemma} [Quisque blandit tempus nunc]
Sed interdum nisl pretium non. Mauris sodales consequat risus vel consectetur. Aliquam erat volutpat. Nunc sed sapien ligula. Proin faucibus sapien luctus nisl feugiat convallis faucibus elit cursus. Nunc vestibulum nunc ac massa pretium pharetra. Nulla facilisis turpis id augue venenatis blandit. Cum sociis natoque penatibus et magnis dis parturient montes, nascetur ridiculus mus.
\end{lemma}

Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.

\section{Morbi eros magna}

Morbi eros magna, vestibulum non posuere non, porta eu quam. Maecenas vitae orci risus, eget imperdiet mauris. Donec massa mauris, pellentesque vel lobortis eu, molestie ac turpis. Sed condimentum convallis dolor, a dignissim est ultrices eu. Donec consectetur volutpat eros, et ornare dui ultricies id. Vivamus eu augue eget dolor euismod ultrices et sit amet nisi. Vivamus malesuada leo ac leo ullamcorper tempor. Donec justo mi, tempor vitae aliquet non, faucibus eu lacus. Donec dictum gravida neque, non porta turpis imperdiet eget. Curabitur quis euismod ligula.

%%
%% Bibliography
%%

%% Please use bibtex,

\bibliography{lipics-v2021-sample-article}

\appendix

\section{Styles of lists, enumerations, and descriptions}\label{sec:itemStyles}

List of different predefined enumeration styles:

\begin{itemize}
\item \verb|\begin{itemize}...\end{itemize}|
\item \dots
\item \dots
%\item \dots
\end{itemize}

\begin{enumerate}
\item \verb|\begin{enumerate}...\end{enumerate}|
\item \dots
\item \dots
%\item \dots
\end{enumerate}

\begin{alphaenumerate}
\item \verb|\begin{alphaenumerate}...\end{alphaenumerate}|
\item \dots
\item \dots
%\item \dots
\end{alphaenumerate}

\begin{romanenumerate}
\item \verb|\begin{romanenumerate}...\end{romanenumerate}|
\item \dots
\item \dots
%\item \dots
\end{romanenumerate}

\begin{bracketenumerate}
\item \verb|\begin{bracketenumerate}...\end{bracketenumerate}|
\item \dots
\item \dots
%\item \dots
\end{bracketenumerate}

\begin{description}
\item[Description 1] \verb|\begin{description} \item[Description 1]  ...\end{description}|
\item[Description 2] Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\item[Description 3]  \dots
%\item \dots
\end{description}

\cref{testenv-proposition} and \autoref{testenv-proposition} ...

\section{Theorem-like environments}\label{sec:theorem-environments}

List of different predefined enumeration styles:

\begin{theorem}\label{testenv-theorem}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{theorem}

\begin{lemma}\label{testenv-lemma}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{lemma}

\begin{corollary}\label{testenv-corollary}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{corollary}

\begin{proposition}\label{testenv-proposition}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{proposition}

\begin{conjecture}\label{testenv-conjecture}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{conjecture}

\begin{observation}\label{testenv-observation}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{observation}

\begin{exercise}\label{testenv-exercise}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{exercise}

\begin{definition}\label{testenv-definition}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{definition}

\begin{example}\label{testenv-example}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{example}

\begin{note}\label{testenv-note}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{note}

\begin{note*}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{note*}

\begin{remark}\label{testenv-remark}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{remark}

\begin{remark*}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{remark*}

\begin{claim}\label{testenv-claim}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{claim}

\begin{claim*}\label{testenv-claim2}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{claim*}

\begin{proof}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{proof}

\begin{claimproof}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{claimproof}

\end{document}

### 2025-02-14 paper.tex
Motivation - type providers, AI assistants

@@
 
+\newcommand{\ident}[1]{\textsf{#1}}
+
 \bibliographystyle{plainurl}
-\title{The Choose-Your-Own-Adventure Calculus} % (Pearl)
+\title{The Choose-Your-Own-Adventure Calculus (Pearl/Brave New Idea)}
 
@@
 
-The main contribution of this paper is conceptual rather than technical. Programming language
-researchers have long focused on \emph{programming languages} rather than on a more general notion
-of interactive \emph{programming systems} \cite{rpg-2012-revolution,jakubovic-2023-techdims}.
-However, programming language theory methods can be equally useful for studying interactive
-systems~\cite{adams-2025-grove}. The choose-your-own-adventure-calculus is another step in this
-direction. It brings to light a pattern that is perhaps not surprising, but is very easy to
-overlook until it is given a name.
+The main contribution of this paper is conceptual rather than technical. We capture a
+pattern that is perhaps not surprising in retrospect, but that is easy to overlook until it is
+given a name. We use formal programming language theory methods to precisely describe
+interesting aspects of the pattern. Moreover, our work also confirms that programming
+language theory methods can be extremely effective for studying not just \emph{programming
+languages}, but also interactive \emph{programming systems}~\cite{jakubovic-2023-techdims}.
 
@@
 
-~
+Computer scientists studying programming have long focused on programming languages as syntactic
+entities, sometimes neglecting the interactive environments in which the languages are inevitably
+embedded \cite{rpg-2012-revolution}. Notably, in many of the motivating examples that we draw
+from in this section, the interactive aspect of the system is only described in supplementary
+materials such as tutorial notes \cite{brady-2015-idris}, workshop papers \cite{syme-2013-inforich}
+or user guides \cite{altenkirch-1994-alf}. Only recently, programming language theory has been used
+to look at interactive environments \cite{adams-2025-grove,mayer-2018-bidirectional}. Our work aims
+to contribute to this research direction.
+
+The following four sections briefly review four different instances of the choose-your-own-adventure
+interaction pattern. In all of those, an interactive editor offers the user some kind of a
+completion list during working with the system.
+
+\begin{figure}[t]
+\includegraphics[width=0.7\textwidth]{fig/worldbank.png}
+\caption{F\# code editor showing completions offered by the World Bank type provider.}
+\label{fig:worldbank}
+\end{figure}
+
+\subparagraph{Type providers}
+
+F\# type providers \cite{syme-2013-inforich} are a mechanism that make it possible to integrate
+external data sources into the F\# type system. A type provider is a compiler extension, loaded and
+executed at compile-time and at edit-time. It can run arbitrary code to read the structure of
+external data and use it to generate a suitable statically-typed representation of the external
+data, typically as objects with members. Type providers can, for example, infer the type from a
+sample JSON \cite{petricek-2016-fsdata} or read a database schema.
+
+The example in Figure~\ref{fig:worldbank} shows a simple type provider for accessing information
+from the World Development Indicators database. The provided \ident{wb} object allows the programmer
+to access any indicator of any country in the database by choosing an appropriate \ident{[Country]}
+and an \ident{[Indicator]} in a chain of members
+\ident{wb}.\ident{Countries}.\ident{[Country]}.\ident{Indicator}.\ident{[Indicator]}.
+The result is a time series with values of the given indicator and country. More generally,
+the example can be seen as a special case of a type provider for slicing n-dimensional
+data cube \cite{petricek-2022-thegamma}. Here, we choose a fixed value for two of the three
+dimensions (country, indicator, time).
+
+When using the type provider, the user types the first line of code and trigger auto-completion
+by typing \ident{wb} followed by the dot. The rest of the code is constructed by choosing an
+option from a list and typing another dot.\footnote{This interaction pattern has been
+lightheartedly called \emph{dot-driven development} by Phil Trelford \cite{seemann-2021-head}.}
+
+\newpage
+
+\begin{figure}[t]
+\includegraphics[width=0.48\textwidth]{fig/thegamma1.png}~~
+\includegraphics[width=0.48\textwidth]{fig/thegamma2.png}
+\caption{Constructing a query in The Gamma. We count the number of gold and silver medals for
+each athlete and sort the data by the number of gold medals.}
+\label{fig:thegamma}
+\end{figure}
+
+\subparagraph{Data exploration}
+
+The Gamma~\cite{petricek-2022-thegamma} is a web-based programmatic data exploration environment
+for non-programmers. In The Gamma, type providers are the primary programming mechanism and they
+are used not just for data access, but also for constructing queries.
+
+The type provider shown in Figure~\ref{fig:thegamma} lets the user construct an SQL-like query by
+repeatedly choosing operations and their parameters \cite{petricek-2017-dotdriven}. It keeps track
+of the data schema and uses the schema to generate all possible valid parameters. When sorting
+data, it generates an object with two members for each columns -- one for ascending and one for
+descending sort. Similarly, grouping first offers all columns as possible sorting keys and then
+lets the user choose from a range of pre-defined aggregations (sum, count, average, concatenate).
+The system evaluates the query on the fly, providing a live preview during editing \cite{petricek-2020-live}.
+
+The interaction pattern is the same as above. The user triggers auto-completion by entering a
+data source and typing dot. They can then repeatedly select an operation and its parameters to
+construct a query. One notable difference is that the structure of the generated types is
+potentially infinite (the user can keep adding further operations) and so the types are generated
+lazily.
+
+\subparagraph {AI assistants}
+
+The third instance of the choose-your-own-adventure interaction pattern comes from work on
+semi-automatic data wrangling tools known as AI assistants \cite{petricek-2023-aias}.
+An AI assistant guides the analyst through a data wrangling problem such as reconciling mismatched
+datasets, filling missing values or inferring data format and types. An AI assistant tries to solve
+the problem automatically and suggests an initial data transformation, but it also generates a
+number of constraints that the user can choose from to refine the initial solution. If the initial
+solution is not correct, the user chooses a constraint and the AI assistant runs again, suggesting
+a new data transformation that respects the constraint.
+
+Figure~\ref{fig:aia} shows an example. It uses the datadiff \cite{sutton-2018-datadiff} AI
+assistant, running in a Wrattler notebook \cite{petricek-2018-wrattler}, to merge broadband
+quality data published by Ofcom for two subsequent years. The format of the CSV files differs.
+Some columns were added, some removed, all were renamed and their order has changed. In the
+example, we selected 6 columns from the year 2015 and want to find matching data from 2014.
+
+When the AI assistants runs automatically, it incorrectly maps the \ident{Urban.rural} (2014)
+column to \ident{Nation} (2015). This happens because both columns are categorical and have three
+values with similar distribution, but a data analyst can easily spot the mistake. To correct the
+error, they click the ``+'' button to add a constraint and choose \ident{Don't match Urban.rural
+and Nation} to specify that the two columns should not be matched. Datadiff then runs again and
+finds the correct matching.
+
+\begin{figure}[t]
+  \includegraphics[width=0.9\textwidth]{fig/aia.png}
+  \caption{aia}
+  \label{fig:aia}
+\end{figure}
+
+The interaction patter is the same as in the previous two cases. The analyst constructs the
+correct data transformation by repeatedly choosing from a list of options, until they obtain
+the desired result. However, the way the interaction pattern is used in this case differs.
+First, in the case of type providers, we are constructing a program (adding operations to a
+method chain). Now, we are collecting constraints and the AI assistant then synthesizes a
+data transformation (a script) based on those constraints. Second, in the case of type providers,
+the completion list offered all possible members of the object. Now, the list offers a choice of
+constraints that are recommended by the AI assistant, which may not be complete.
+
+\subparagraph{Interactive theorem provers}
+
+yo
+
+\newpage
 
@@
 
-\cite{jakubovic-2023-techdims}
-\cite{petricek-2020-live}
+\cite{rein-2019-exploratory}
 

### 2025-02-15 paper.tex
Theorem provers, formal model

--- 2025-02-13 paper.tex
@@
 
+\newcommand{\ident}[1]{\textsf{#1}}
+\newcommand{\sigmaN}{\sigma_0}
+\newcommand{\select}{\textnormal{\ident{choose}}}
+\newcommand{\preview}{\textnormal{\ident{preview}}}
+\newcommand{\links}{\textnormal{\ident{links}}}
+\newcommand{\choices}{\textnormal{\ident{choices}}}
+\newcommand{\op}{\textit{op}}
+\newcommand{\argmax}{\textit{arg}\,\textit{max}\,}
+
 \bibliographystyle{plainurl}
-\title{The Choose-Your-Own-Adventure Calculus} % (Pearl)
+\title{The Choose-Your-Own-Adventure Calculus (Pearl/Brave New Idea)}
 
@@
 
-Some of the most remarkable results in mathematics reveal surprising connections between different
+Some of the most remarkable results in mathematics reveal connections between different
 branches of the discipline. The aim of this paper is to point out a modest, but still
@@
 To distinguish the different ways of embedding the interactions in the edited programs, we
-also formally define \emph{base-level} and \emph{meta-level} integration of the system.
+also formally define \emph{internal} and \emph{external} mode of system integration.
 
@@
 
-The main contribution of this paper is conceptual rather than technical. Programming language
-researchers have long focused on \emph{programming languages} rather than on a more general notion
-of interactive \emph{programming systems} \cite{rpg-2012-revolution,jakubovic-2023-techdims}.
-However, programming language theory methods can be equally useful for studying interactive
-systems~\cite{adams-2025-grove}. The choose-your-own-adventure-calculus is another step in this
-direction. It brings to light a pattern that is perhaps not surprising, but is very easy to
-overlook until it is given a name.
+The main contribution of this paper is conceptual rather than technical. We capture a
+pattern that is perhaps not surprising in retrospect, but that is easy to overlook until it is
+given a name. We use formal programming language theory methods to precisely describe
+interesting aspects of the pattern. Moreover, our work also confirms that programming
+language theory methods can be extremely effective for studying not just \emph{programming
+languages}, but also interactive \emph{programming systems}~\cite{jakubovic-2023-techdims}.
 
 \newpage
+
+\begin{figure}[t]
+  \includegraphics[width=0.7\textwidth]{fig/worldbank.png}
+  \caption{F\# code editor showing completions offered by the World Bank type provider.}
+  \label{fig:worldbank}
+\end{figure}
 
@@
 
+Computer scientists studying programming have long focused on programming languages as syntactic
+entities, sometimes neglecting the interactive environments in which they are inevitably
+embedded \cite{rpg-2012-revolution}. Notably, in many of the motivating examples that we draw
+from in this section, the interactive aspect of the system is only described in supplementary
+materials \cite{brady-2015-idris,syme-2013-inforich,altenkirch-1994-alf}. Only recently, programming
+language theory started to be used to study interactive environments
+\cite{adams-2025-grove,mayer-2018-bidirectional}. Our work contributes to this research direction.
+
+The following sections review four different instances of the choose-your-own-adventure
+interaction pattern. In all of those, an interactive editor offers the user some kind of~a~com\-pletion list during working with the system.
+
+\subparagraph{Type providers}
+
+F\# type providers \cite{syme-2013-inforich} are a mechanism for integrating external data
+sources into the F\# type system. A type provider is a compiler extension, loaded and
+executed at compile-time and at edit-time. It can run arbitrary code to read the structure of
+external data and use it to generate a suitable statically-typed representation of the
+data, typically as objects with members. Type providers can, for example, infer the type from a
+sample JSON \cite{petricek-2016-fsdata} or read a database schema.
+
+The example in Figure~\ref{fig:worldbank} shows a simple type provider for accessing information
+from the World Development Indicators database. The provided \ident{wb} object allows the programmer
+to access any indicator of any country in the database by choosing an appropriate \ident{[Country]}
+and an \ident{[Indicator]} in a chain of members
+\ident{wb}.\ident{Countries}.\ident{[Country]}.\ident{Indicator}.\ident{[Indicator]}.
+The result is a time series with values for the given indicator and a country. More generally,
+the example can be seen as a special case of a type provider for slicing n-dimensional
+data cube \cite{petricek-2022-thegamma} -- we choose a fixed value for two of the three
+dimensions (country, indicator, time).
+
+When using the type provider, the user types the first line of code and triggers auto-completion
+by typing \ident{wb} followed by the dot. The rest of the code is constructed by choosing an
+option from a list and typing another dot.\footnote{This interaction pattern has been
+lightheartedly called \emph{dot-driven development} by Phil Trelford \cite{seemann-2021-head}.}
+
+\newpage
+
+\begin{figure}[t]
+\includegraphics[width=0.48\textwidth]{fig/thegamma1.png}~~
+\includegraphics[width=0.48\textwidth]{fig/thegamma2.png}
+\caption{Constructing a query in The Gamma. We count the number of gold and silver medals for
+each athlete and sort the data by the number of gold medals.}
+\label{fig:thegamma}
+\end{figure}
+
+\subparagraph{Data exploration}
+
+The Gamma~\cite{petricek-2022-thegamma} is a programmatic data exploration environment
+for non-programmers. In The Gamma, type providers are the primary programming mechanism. They
+are used not just for data access, but also for constructing queries.
+
+The type provider shown in Figure~\ref{fig:thegamma} lets the user construct an SQL-like query by
+repeatedly choosing operations and their parameters \cite{petricek-2017-dotdriven}. It keeps track
+of the schema and uses it to generate all possible valid parameters. When sortin data, it generates
+an object with two members for each columns -- one for ascending and one for descending sort.
+Similarly, the grouping operation first offers all columns as possible grouping keys and then
+lets the user choose from a range of pre-defined aggregations (sum, count, average, concatenate).
+The system also evaluates the query on the fly, providing a live preview during editing \cite{petricek-2020-live}.
+
+The interaction pattern is the same as before. After the usertriggers auto-completion, they
+repeatedly select an operation and its parameters to construct a query. One notable difference
+is that the structure of the generated types is potentially infinite (the user can keep adding
+further operations) and so the types are generated lazily.
+
+\subparagraph{AI assistants}
+
+The third instance of the choose-your-own-adventure interaction pattern comes from the work on
+semi-automatic data wrangling tools known as AI assistants \cite{petricek-2023-aias}.
+An AI assistant guides the analyst through a data wrangling problem such as reconciling mismatched
+datasets, filling missing values or inferring data format and types. An AI assistant solves
+the problem automatically and suggests an initial data transformation, but it also generates a
+number of constraints that the user can choose from to refine the initial solution. If the initial
+solution is not correct, the user chooses a constraint and the AI assistant runs again, suggesting
+a new data transformation that respects the constraint.
+
+Figure~\ref{fig:aia} shows an example. It uses the datadiff \cite{sutton-2018-datadiff} AI
+assistant, running in a Wrattler notebook \cite{petricek-2018-wrattler}, to merge broadband
+quality data published by Ofcom for two subsequent years. The format of the CSV files for the
+two years differs. Columns were added, removed, renamed and their order has changed. In the
+example, we selected 6 columns from the year 2015 and want to find matching data from 2014.
+
+When the AI assistants runs automatically, it correctly maps the numerical columns, but it
+incorrectly maps the \ident{Urban.rural} (2014) column to \ident{Nation} (2015). This happens
+because both columns are categorical and have three values with similar distribution. A data
+analyst can easily spot the mistake. They click the ``+'' button to add a constraint and choose
+\ident{Don't match Urban.rural and Nation} to specify that the two columns should not be matched.
+Datadiff then runs again and finds the correct matching.
+
+\begin{figure}[t]
+  \includegraphics[width=0.9\textwidth]{fig/aia.png}
+  \caption{Using the datadiff AI assistant to reconcile the structure of the two datasets.
+    The user is offered a list of constraints to prevent or force matching between specific columns.}
+  \label{fig:aia}
+\end{figure}
+
 ~
+
+The interaction patter is the same as in the previous two cases. The analyst constructs the
+correct data transformation by repeatedly choosing from a list of options, until they obtain
+the desired result. However, the way the interaction pattern is implemented differs.
+First, in the case of type providers, we are gradually constructing a program by adding operations to
+a method chain. Now, the AI assistant synthesizes a data transformation (program) and we are
+gradually adding constraints to control the synthesis. Second, in the case of type providers,
+the completion list offered all possible members of the object. Now, the list offers
+constraints recommended by the AI assistant which may not be complete.
+
+\newpage
+
+\begin{figure}[t]
+  \includegraphics[width=0.6\textwidth]{fig/alf.png}
+  \caption{Constructing a proof of the transitivity of the $\leq$ relation in the \textsc{Alf}
+    editor. The user is offered a range of variables and constructors in scope at the current
+    location. \cite{altenkirch-1994-alf}}
+  \label{fig:alf}
+\end{figure}
+
+\subparagraph{Interactive theorem provers}
+A fourth example of the choose-your-own-adventure interactive pattern can be found in interactive
+theorem provers. When writing programs in systems like Idris \cite{brady-2021-idris2}, the
+user typically works by stating the desired conclusion and filling the implementation with a hole.
+The system provides a range of interactive editing capabilities to fill the holes \cite{mcbride-1999-dependently}.
+It can, for example, generate a case split or search for a proof \cite{brady-2015-idris}.
+
+Systems like Idris provide key bindings to invoke the completions, but the functionality could
+also be offered through a user interface. An example that illustrates this is the interactive editor
+for the \textsc{Alf} theorem prover \cite{magnusson-1994-alf}, which is based on the refinement of
+an incomplete proof object \cite{altenkirch-1994-alf}. This is illustrated in Figure~\ref{fig:alf}.
+The user is proving the transitivity of the $\leq$ relation for Peano arithmetic natural numbers.
+They pattern match on the proof argument $p$ and complete the first branch. For the second branch,
+they need to fill a hole $?_{p2}$ (called a wildcard in \textsc{Alf}). They trigger a completion
+and a pop-up menu shows the available variables and constructors, including \ident{leq\_trans} that
+can be used to complete the proof. After choosing \ident{leq\_trans}, two new holes are generated
+for its arguments. Those can be, again, filled interactively, by choosing $p_1$ and $p$ from the
+completion.
+
+The interaction pattern is again the same. The user repeatedly triggers a completion and uses
+it to refine and complete their proof by filling holes. There are subtle differences too. Unlike
+with AI assistants, each completion directly refines the proof that the user is editing. Unlike
+with type providers, a completion may generate multiple new holes, rather than just adding to a
+chain of operations.
+
+The \textsc{Alf} editor is a historical example, but a similar user interface could be built for
+systems like Idris or Coq. The two would work differently. As in \textsc{Alf}, Idris source code
+represents the proof itself and a completion would replace a hole with a suggested term. In Coq,
+the proof is a series of tactic invocations and so selected completions would be added to this
+list and would form a trace of the interaction with the user.
+
+\newpage
 
@@
 
-~
+A system that implements the choose-your-own-adventure interaction pattern repeatedly offers
+the user a range of options to choose from. Each of the options is designated by an identifier.
+The system also maintains a state during the process which determines subsequent options. The state
+may not be visible to the user, but the user can always explicitly request the program constructed
+so far.
+
+We can think of the interaction with the system as navigating through a tree structure, starting
+from a root and choosing one of the possible branches in each step.\footnote{Serving as another
+evidence for the surprising effectiveness of the concept of a tree \cite{nesetril-2005-strom}.}
+In the following definition, the key $\choices$ operation can thus be seen as returning
+branches of a given node.
+
+\begin{definition}[Choose-your-own-adventure system]\label{def:calculus}
+Given expressions $e\in \mathbb{E}$ and states $\sigma \in \Sigma$, a choose-your-own-adventure
+system is a pair of operations \choices, \select\ such that:
+
+\vspace{-0.25em}
+\begin{itemize}
+  \item $\choices(\sigma) = (\iota_1\mapsto\sigma_1, \ldots, \iota_n\mapsto\sigma_n)$ is
+    an operation that takes a state and generates options designated by an identifier $\iota_i$
+    and represented by a state $\sigma_i$,
+  \item $\select(\sigma) &=& e$ is an operation that returns generated program for a given state.
+\end{itemize}
+\end{definition}
+
+The definition is not a programming language calculus in the usual sense in that it does not
+define a concrete syntax with reduction rules. It is an abstract algebraic structure that captures
+the structure of a system that supports the choose-your-own-adventure interaction pattern.
+The definition is close to that of an AI assistant \cite{petricek-2023-aias}, which is written
+using a language specific for the data wrangling domain (such as cleaning scripts or input and
+output data) but is structurally similar. It is also worth noting that the definition may describe
+not just trees, but also graphs with cycles -- a system can return to an already visited state.
+This is not practically useful, but it does not pose a theoretical problem.
+
+\subparagraph{External mode of embedding}
+One of the subtle questions about the choose-your-own-adventure pattern raised in the introduction
+concerns the different ways in which a trace of the interaction is embedded in the interactively
+constructed program. In the \emph{external mode}, the interaction results in code that becomes a
+part of the edited program, but it is not possible to reconstruct the steps used to generate the code.
+
+The choose-your-own-adventure interaction pattern is typically used to complete a partial program.
+To model this, we assume that the host language has a notion of a hole, written as~$?$
+and that a user can select a part of program to invoke the completion on. We write $E[e]$ for a
+completion context, akin to evaluation contexts in operational semantics.
+
+We assume that, for a program containing a hole in a completion context $E[?]$, we can construct
+an initial choose-your-own-adventure state using an operation
+$\ident{init}(E[?])=\sigmaN$.
+
+\begin{definition}\label{def:external}
+An expression $E[?]$ is completed as $E[e]$ via external embedding of an interaction with
+a choose-your-own-adventure system consisting of \choices\ and \select\ if:
+
+\vspace{-0.25em}
+\raggedright
+\begin{enumerate}
+\item $\ident{init}(E[?]) = \sigmaN$ obtains the initial state of a choose-your-own-interaction system,
+\item $\sigma_n$ is a system state such that $\forall i\in{1\ldots n}.(\iota_i\mapsto\sigma_i)\in\choices{(\sigma_{n-1})}$,
+  i.e.,~the user\\ makes a series of choices resulting in a final state of the system $\sigma_n$,
+\item $E[e]$ where $e=\select(\sigma_n)$, i.e.,~the final program is constructed by replacing the hole
+  in the completion context with the expression $e$ generated from $\sigma_n$.
+\end{enumerate}
+\end{definition}
+
+As we will see when we revisit the earlier examples formally, the external mode of embedding is
+used, for example, in the case of interactive theorem provers like \textsc{Alf} or Idris.
+In those systems, the user triggers the completion on a proof (program) containing a hole.
+They then fill the hole and, possibly iteratively, further holes in the generated proof. The final
+expression is embedded in the source code, but it does not indicate what options, identified by
+$\iota_1, \ldots, \iota_n$, were selected in the process.
+
+\subparagraph{Internal mode of embedding}
+In the \emph{internal mode}, a trace of the interaction with a choose-your-own-adventure system
+is embedded directly in the constructed program. This is the case with type providers, where a
+user chooses a sequence of object members to be accessed. The same would be the case in a completion
+system for Coq that would offer tactics to apply, becuase the resulting proof would contain a
+record of the selected tactics.
+
+To talk about the internal mode formally, we again need the \ident{init} operation, but also an
+operation \ident{decode} that extracts identifiers of invoked completions from an expression.
+An internal embedding is the same as external embedding with an additional constraint:
+
+\begin{definition}\label{def:internal}
+An expression $E[?]$ is completed as $E[e]$ via internal embedding of an interaction with
+a choose-your-own-adventure system consisting of \choices\ and \select\ if:
+
+\vspace{-0.25em}
+\raggedright
+\begin{enumerate}
+\item $E[?]$ is completed as $E[e]$ via external embedding using \ident{init}, \choices\ and \select\\
+  through a series of choices designated by identifiers $\iota_1, \ldots, \iota_n$,
+\item it also holds that $\ident{decode}(e)=(\iota_1, \ldots, \iota_n)$.
+\end{enumerate}
+\end{definition}
+
+If a choose-your-own-adventure system is integrated in a programming language
+through internal embedding, we can reconstruct the choices through which the user constructed
+an expression $e$ in a completion context $E[e]$, assuming they used the interactive system rather
+than entering the code directly. This also means that we can reconstruct the final state $\sigma_n$
+of the system by starting from $\ident{init}(E[?])$ and following the choices specified by
+$\iota_1, \ldots, \iota_n$.
 
@@
 ~
+
+The internal mode can also be used when
+working with theorem provers.
+
+In Coq, where proofs are represented as series of tactics,
+
+However, this would also be case in a
+completion engine for the Coq proof system that would offer a choice of tactics to apply, because
+Coq proofs are represented
+
+\subsection{Data exploration}
+
+~
+
+\subsection{AI assistants}
+
+~
+
+\subsection{Interactive theorem proving}
+
+~
+
+\newpage
 

### 2025-02-16 paper.tex
The Gamma example

@@
 \newcommand{\ident}[1]{\textsf{#1}}
+\newcommand{\ddident}[1]{\guillemotleft\ident{#1}\guillemotright}
 \newcommand{\sigmaN}{\sigma_0}
 \newcommand{\select}{\textnormal{\ident{choose}}}
-\newcommand{\preview}{\textnormal{\ident{preview}}}
-\newcommand{\links}{\textnormal{\ident{links}}}
 \newcommand{\choices}{\textnormal{\ident{choices}}}
 \newcommand{\op}{\textit{op}}
-\newcommand{\argmax}{\textit{arg}\,\textit{max}\,}
+\newcommand{\vect}[1]{\lbrack #1 \rbrack}
+%\newcommand{\preview}{\textnormal{\ident{preview}}}
+%\newcommand{\links}{\textnormal{\ident{links}}}
+%\newcommand{\argmax}{\textit{arg}\,\textit{max}\,}
 
@@
 
-\subparagraph{Type providers}
+\subparagraph{Type providers.}
 
@@
 
-\subparagraph{Data exploration}
+\subparagraph{Data exploration.}
 
@@
 
-\subparagraph{AI assistants}
+\subparagraph{AI assistants.}
 
@@
 
-\subparagraph{Interactive theorem provers}
+\subparagraph{Interactive theorem provers.}
 A fourth example of the choose-your-own-adventure interactive pattern can be found in interactive
@@
     and represented by a state $\sigma_i$,
-  \item $\select(\sigma) &=& e$ is an operation that returns generated program for a given state.
+  \item $\select(\sigma) = e$ is an operation that returns generated program for a given state.
 \end{itemize}
@@
 
-\subparagraph{External mode of embedding}
+\subparagraph{External mode of embedding.}
 One of the subtle questions about the choose-your-own-adventure pattern raised in the introduction
@@
 
-\subparagraph{Internal mode of embedding}
+\subparagraph{Internal mode of embedding.}
 In the \emph{internal mode}, a trace of the interaction with a choose-your-own-adventure system
@@
 
+\newpage
 \section{Examples}
@@
 
+We now revisit the four examples from Section~\ref{sec:motivation} and show how they fit the
+above formal model. All four examples rely on some domain-specific logic. We describe what
+information the logic provides, but do not model it formally. This has been done elsewhere,
+in works describing the individual systems.
+
+To show how the model lets us distinguish subtle details of interactive programming systems,
+we start with a model of data exploration system that is inspired by The Gamma, but differs in
+one notable way. We then discuss type providers more generally and show how to correctly model
+The Gamma. We then revisit the remaining two examples.
+
+\subsection{Data exploration}
+
+In The Gamma, the choose-your-own-adventure interaction pattern is used to construct a
+query that transforms the given input data. The query is a sequence of operations with parameters,
+$\op(p_1, \ldots, p_n)$, loosely modelled after relational algebra \cite{codd-1970-relational}.
+
+In The Gamma, the query is hidden from the data analyst. Behind the scenes, the system generates
+objects with members and the identifiers designating individual options are the names of those
+members. The operation is encapsulated in the code of the accessor of the member. In the
+simplified model in this section we ignore this fact. The model presented here directly
+generates code that calls the underlying operations. For example, assume that the user makes
+the following choices:
+\[
+\textnormal{\ddident{group data}\,.\,\ddident{by Athlete}\,.\,\ddident{sum Gold}\,.\,\ddident{count all}\,.\,\ddident{then}}
+\]
+In The Gamma, the individual identifiers become object members and they are included as
+a member chain in the generated code. In the following simplified model, the completion
+instead fills the hole with an expression representing the operation:
+\[
+\ident{group}(\texttt{"Athlete"}, \ident{sum}(\texttt{"Gold"}), \ident{count}())
+\]
+The two approaches have different human-computer interaction trade-offs. In terms of cognitive
+dimensions \cite{green-1989-cogdims,blackwell-2003-cogdims}, the latter has a greater closeness
+of mapping, while the former is less cognitively demanding to read for a non-programmer.
+As discussed in Section~\ref{sec:properties}, the two implementations of the choose-your-own-adventure
+interaction pattern also differ in terms of their formal properties.
+
+\subparagraph{Formal model.}
+The options generated by The Gamma let the user select both the next operation and the parameters
+of the previously selected operation. The available operations and parameters are generated
+based on a schema $S$ that is transformed by the operations. The state of the system $\sigma$
+contains the current schema $S$ and the operations applied so far. In the following, we write
+$\op(\boldsymbol{p})$ for an operation with a vector of parameters:
+\[
+\sigma = S, \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n})}\\
+\]
+The behaviour of the $\choices$ operation depends on whether the last operation in the sequence
+expects further parameters or whether it is fully-specified. In the first case, the recommendation
+engine generates possible additional parameter values $p_', p_'', \ldots$ based on the schema $S$,
+the operation $\op_n$ and the already known parameters $\boldsymbol{p_n}$.
+The $\choices$ operation then generates options that add the additional parameter. We generated the
+identifiers $\iota_',\iota_'',\ldots$ based on the state and the parameter value, such as
+\ddident{by Gold descending}. Note that adding a parameter may also result in a new schema
+$S', S'', \ldots$ (which the recommendation engine computes based on the previous schema and the new parameter):
+\[
+\begin{array}{l}
+\choices(S, \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n})}) =\\
+\qquad (S', \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n},p')}), \\
+\qquad (S'', \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n},p'')}), ~\ldots
+\end{array}
+\]
+If the last operation takes no further parameters, the system produces a choice
+of possible next operations $\op', \op'', \ldots$. Again, we are also given new schemas $S', S'', \ldots$
+and we generate identifiers $\iota',\iota'',\ldots$ based on the operation name. The $\ident{choices}$
+operation then returns options that add the additional operation:
+\[
+\begin{array}{l}
+\choices(S, (\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n}))) =\\
+\qquad (S', \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n}), \op'()}), \\
+\qquad (S'', \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n}), \op''()}), ~\ldots
+\end{array}
+\]
+Finally, the $\select$ operation takes the state $\sigma$ and generates an expression that represents
+the data transformation. This is only possible if all parameters are fully-specified. For simplicity,
+assume that $k$ is the index of the last fully-specified operation (either $n$ or $n-1$). If the
+host language lets us compose functions using $f\circ g$, we can write:
+\[
+\select(S, (\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n}))) = \op_1(\boldsymbol{p_1}) \circ \ldots \circ \op_k(\boldsymbol{p_k})
+\]
+The recommendation engine behind The Gamma provides a domain-specific logic for generating
+possible operations and their parameters based on the current schema of the data. As the above
+definition shows, this underlying engine can be easily exposed through the common
+choose-your-own-adventure interface.
+
+\subsection{Type providers}
+
+revisit
+
+~
+
+\subsection{AI assistants}
+
+~
+
+\subsection{Interactive theorem proving}
 ~
@@
 Coq proofs are represented
-
-\subsection{Data exploration}
-
-~
-
-\subsection{AI assistants}
-
-~
-
-\subsection{Interactive theorem proving}
 

### 2025-02-17 paper.tex
TPs and AIAs examples

@@
 %\newcommand{\links}{\textnormal{\ident{links}}}
-%\newcommand{\argmax}{\textit{arg}\,\textit{max}\,}
+\DeclareMathOperator{\argmax}{arg\,max}
 
@@
 \vspace{-0.25em}
+\raggedright
 \begin{itemize}
-  \item $\choices(\sigma) = (\iota_1\mapsto\sigma_1, \ldots, \iota_n\mapsto\sigma_n)$ is
-    an operation that takes a state and generates options designated by an identifier $\iota_i$
+  \item $\choices(\sigma) = \{\iota_1\mapsto\sigma_1, \ldots, \iota_n\mapsto\sigma_n\}$ is
+    an operation that takes a state and \\ generates options designated by an identifier $\iota_i$
     and represented by a state $\sigma_i$,
@@
 \choices(S, \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n})}) =\\
-\qquad (S', \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n},p')}), \\
-\qquad (S'', \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n},p'')}), ~\ldots
+\qquad \{\; \iota' \mapsto (S', \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n},p')}), \\
+\qquad\;\;    \iota'' \mapsto (S'', \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n},p'')}), ~\ldots~\}
 \end{array}
@@
 \choices(S, (\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n}))) =\\
-\qquad (S', \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n}), \op'()}), \\
-\qquad (S'', \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n}), \op''()}), ~\ldots
+\qquad \{\; \iota' \mapsto (S', \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n}), \op'()}), \\
+\qquad \;\; \iota'' \mapsto (S'', \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n}), \op''()}), ~\ldots~\}
 \end{array}
@@
 \subsection{Type providers}
+The type provider mechanism in F\# operates at the level of the type system. It is not a merely an
+editor feature. A type provider for a data source, such as the World Development Indicators database,
+generates a collection of types that model the external data source. In F\#, the types are classes
+with members that implement the logic to retrieve data at runtime.
 
-revisit
+The auto-completion mechanism in F\# code editors, which implements the choose-your-own-adventure
+interaction pattern, is not specific to type providers. It offers a list of members of an object
+based on its type. We model the completion as an iterative process, repeatedly adding further
+members to a chain. The state $\sigma$ thus consists of an initial expression on which the
+completion is invoked, a chain of selected members and the type of the last member.
 
-~
+To model the completion mechanism, we also need to model information about types. We loosely
+follow the Foo calculus model \cite{petricek-2016-fsdata} and write $\mathbb{C}$ for a set
+of class definitions, each consisting of an implicit class constructor and a collection of
+members $M$:
+\[
+\begin{array}{rcl}
+\sigma &=& e . \iota_1 . \lbrack\ldots\rbrack . \iota_n, C\\
+\mathbb{C} &=& \{ C \mapsto \ident{type}~C(\overline{x:\tau}) = \overline{M},~\ldots~ \}\\
+M &=& \ident{member}~\iota\!:\!C=e
+\end{array}
+\]
+Each member in the Foo calculus consists of a of a name $\iota$, return type $C$ and implementation
+$e$. For our purposes, we only need the type information and so the operations that define the
+choose-your-own-adventure are parameterized by the set of classes $\mathbb{C}$.
+
+The $\choices_\mathbb{C}$ operation finds the class definition corresponding to the type of the
+last member in the current chain. It offers choices appending each of the available
+members to the current chain. The $\select_\mathbb{C}$ operation returns the constructed
+member chain:
+
+\[
+\begin{array}{l}
+\choices_\mathbb{C}(e . \iota_1 . \lbrack\ldots\rbrack . \iota_n, C) =\\
+\qquad \{\; \iota' \mapsto (e . \iota_1 . \lbrack\ldots\rbrack . \iota_n . \iota', C')\\
+\qquad \;\; \iota'' \mapsto (e . \iota_1 . \lbrack\ldots\rbrack . \iota_n . \iota'', C''), ~\ldots~\}\\[1em]
+\end{array}
+\]\vspace{-2em}
+\[
+\begin{array}{rcl}
+\quad \textnormal{where}~\mathbb{C}(C) &=& \ident{type}~C(\overline{x:\tau}) = M', M'', \ldots\\
+\quad \textnormal{and}~M' &=& \ident{member}~\iota'\!:\!C'=e'
+\end{array}
+\]\vspace{-0.5em}
+\[
+\select_\mathbb{C}(e . \iota_1 . \lbrack\ldots\rbrack . \iota_n, C) = e . \iota_1 . \ldots . \iota_n
+\]
+The model does not directly refer to type providers. Those are responsible solely for generating
+the type definitions in $\mathbb{C}$ as documented in earlier work \cite{petricek-2016-fsdata}.
+It is worth noting that the type provider for data exploration, implemented by The Gamma,
+additionally needs to generate classes lazily \cite{petricek-2017-dotdriven}. To model this aspect,
+the simple lookup $\mathbb{C}(C)$ needs to be replaced with an operation that returns the type
+definition, alongside with a new context $\mathbb{C}'$ that contains additional generated
+type definitions (return types for all the members of the class~$C$).
+
+The model follows the internal mode of embedding the interaction in the program. It is easy to
+define the \ident{decode} operation that takes the resulting generated expression and returns the
+sequence of choices, because the choices are items of the member chain. A slight caveat is that
+the completion is not invoked on an empty hole, but on a hole that contains the initial expression
+on which the completion is applied. We can model this using filled holes \cite{omar-2019-holes}
+and write $?_e$ for a hole containing the initial expression $e$. The $\ident{init}(?_e)$ operation
+then returns $e$ alongside with an empty chain and the type of $e$.
+
+As noted earlier, The Gamma does not embed query expressions directly into the generated code.
+It uses the same model as type providers and generates choices as members of
+types behind the scenes. We return to the differences between the two models in
+Section~\ref{sec:properties}.
 
 \subsection{AI assistants}
+AI assistants guide the analyst through a data wrangling task. They
+generate a data cleaning script, taking into account constraints selected by the user.
+Most AI assistants obtain the script by performing statistical optimization with respect
+to a set of constraints specified by the user. That is, they look for an expression from the
+set of all possible expressions that optimizes some objective function that assigns score to
+the expression with respect to the given input data. Note that AI assistants do not iterate
+over all possible expressions. They use a machine learning method to approximate a solution
+to the problem.
 
-~
+An optimization-based AI assistant \cite{petricek-2023-aias} thus provides another, very
+different, way of implementing the choose-your-own-adventure pattern. The assistant operates
+with respect to some input data $X$ that does not change during the interaction and so we
+parameterize the choose-your-own-adventure calculus operations by the data.
+The input data $X$ can be actual input data or a representative sample and so the AI assistant
+can be use past data to infer a cleaning script that will be used on new inputs.
+
+The state $\sigma$ consists of a set of constraints specified by the user. We write
+$c$ for individual constraints and $\boldsymbol{c}$ for a set of constraints. The initial
+state is an empty set:
+\[
+\begin{array}{rcl}
+\sigma &=& \{ c_1, \ldots, c_n \}\\
+\sigmaN &=& \emptyset
+\end{array}
+\]
+Unlike in the previous examples, the crucial logic of an AI assistants is implemented in the
+$\select$ operation. The operation runs the optimization algorithm to choose the best cleaning
+script for given constraints. Formally, this can be written using the $\argmax$ operator which
+finds an argument (an expression) for which the given function (scoring function) is maximized.
+The user-specified constraints can either restrict the set of possible expressions or influence
+the scoring function. More formally, we assume that:
+
+\begin{itemize}
+\item $E_{\boldsymbol{c}}\subseteq E$ is a set of expressions that respect constraints $\boldsymbol{c}$,
+\item $Q_{\boldsymbol{c}}(X, e)$ is a scoring function with respect to the constraints $\boldsymbol{c}$,
+  which returns the score of an expression $e$, i.e., how good $e$ is at cleaning the data $X$.
+\end{itemize}
+
+\noindent
+For a given set of constraints $\boldsymbol{c}$, the $\select$ operation looks
+for $e\in E_{\boldsymbol{c}}$ with the largest score:
+\[
+\select_X(\boldsymbol{c}) = \argmax_{e \in E_{\boldsymbol{c}}} Q_{\boldsymbol{c}}(X, e)
+\]
+The actual implementation of the optimization uses various machine learning techniques to
+find the optimal expression. In case of datadiff, $X$ is a pair of datasets $X_1, X_2$ to be
+reconciled. The AI assistant uses the Hungarian algorithm \cite{sutton-2018-datadiff} to construct
+a matching of columns from $X_1$ and $X_2$. The generated expression is a sequence of patches
+that can be applied to $X_2$ in order to reconcile its structure with the sturcture of $X_1$.
+The constraints specified by the user restrict the space of possible column matchings and so they
+affect $E_\boldsymbol{c}$. The scoring function $Q_\boldsymbol{c}$ is independent of
+the constraints and computes a sum of distances between the statistical distributions of the
+columns from $X_1$ and a patched version of $X_2$.
+
+The $\choices$ operation is responsible for generating possible constraints that the user may
+want to add to guide the inference. AI assistants typically offer the user options to prevent
+or adapt some aspect of the cleaning logic inferred by the system. For example, if datadiff
+matches two columns, it will offer a constraint to prevent the matching. It also generates
+constraints that let the user force a specific matching.
+
+To implement $\choices_X$, optimization-based AI assistants first call $\select_X(\sigma)$ to
+get the best expression $e$. Based on this, they generate possible constraints $c_1, c_2, \ldots$
+that the user may want to choose from. The identifiers $\iota_1,\iota_2, \ldots$ provide a
+human-readable description of the constraints. Note that this operation is specific to the particular
+AI assistant. The $\choices_X$ operation then offers a list of constraint sets where the additional
+constraint is added to the previously collected set:
+\[
+\choices_X(\boldsymbol{c}) = \{\iota_1 \mapsto \boldsymbol{c}\cup\{ c_1 \}, \iota_2\mapsto\boldsymbol{c}\cup\{ c_2 \}, \ldots\}
+\]
+The integration of an AI assistant, as described here, has to follow the external mode of embedding.
+The interaction with the assistant results in a cleaning script (expression), but there is no way
+of reconstructing the constrains used to guide the optimization. To support internal embedding, the
+$\select$ operation would need to explicitly include the constraints in the resulting expression.
+However, rerunning the $\select$ operation with the same constraints may result in a different
+cleaning script if the machine learning algorithm is probabilistic.
+
+\newpage
 
@@
 \cite{rein-2019-exploratory}
+
+\section{Limitations}
+
+hard to edit - you cannot easily go back
 

### 2025-02-19 paper.tex
Theorem provers/synthesis

@@
 expects further parameters or whether it is fully-specified. In the first case, the recommendation
-engine generates possible additional parameter values $p_', p_'', \ldots$ based on the schema $S$,
+engine generates possible additional parameter values $p', p'', \ldots$ based on the schema $S$,
 the operation $\op_n$ and the already known parameters $\boldsymbol{p_n}$.
 The $\choices$ operation then generates options that add the additional parameter. We generated the
-identifiers $\iota_',\iota_'',\ldots$ based on the state and the parameter value, such as
+identifiers $\iota',\iota'',\ldots$ based on the state and the parameter value, such as
 \ddident{by Gold descending}. Note that adding a parameter may also result in a new schema
@@
 The constraints specified by the user restrict the space of possible column matchings and so they
-affect $E_\boldsymbol{c}$. The scoring function $Q_\boldsymbol{c}$ is independent of
+affect $E_{\boldsymbol{c}}$. The scoring function $Q_{\boldsymbol{c}}$ is independent of
 the constraints and computes a sum of distances between the statistical distributions of the
@@
 
-\subsection{Interactive theorem proving}
-~
-
-The internal mode can also be used when
-working with theorem provers.
-
-In Coq, where proofs are represented as series of tactics,
-
-However, this would also be case in a
-completion engine for the Coq proof system that would offer a choice of tactics to apply, because
-Coq proofs are represented
-
-~
+\subsection{Theorem proving}
+
+In the previous three sections, we showed how existing formally well-documented systems fit the
+choose-your-own-adventure interaction pattern. Although interactive theorem provers and editors
+for dependently typed languages implement similar kinds of interactions, there is no well-documented
+system that exactly fits the pattern. The closest example is perhaps the recently envisioned
+mixed-mode interaction theorem prover \cite{verter-2024-mixed}. Rather than reframing the
+implementation of an existing system, this section thus outlines a possible implementation.
+
+\subparagraph{Theorem provers.}
+There are two approaches to interacting with an interactive theorem prover. In Coq, the user
+writes a sequence of tactics that transform proof goals. In Agda or Idris, the user writes a
+term, or program, of a type that represents the theorem. Interactive editors exist for both
+types of systems. For Coq, the \texttt{Company-Coq} \cite{pitclaudel-2016-companycoq,aspinall-2000-general}
+extension offers auto-completion, which recommends available tactics, hypotheses and local
+definitions, but it does not filter them based on what is valid in a given context.
+For Idris, the interactive editor \cite{brady-2015-idris} offers a range of commands that
+transform the selected term by adding a case split, a missing case or by automatically searching
+for a proof. In Idris, the system produces valid completions, but those cover only a small number
+of situations.
+
+Despite the different ways of working, an implementation of the choose-your-own-adventure
+pattern for both types of systems would be similar. Based on the sub-goal that the user is
+currently proving, the system would recommend a range of tactics that can be applied to the
+sub-goal. In the case of Coq, the selected tactic would be added to the sequence. In Idris, the
+selected tactic would be applied to transform the current term. The difference is in the mode of
+embedding. A system for Coq would provide internal embedding in that the selected option is added
+to the proof source code. (Much like selecting a completion when using type providers appends a
+member access.) A system for Idris-like language would use the tactic to transform the term,
+making it impossible to reconstruct the sequence of applied tactics as in the external mode of
+embedding.
+
+\begin{figure}
+\[
+\begin{array}{lcl}
+\boxed{\Gamma \vdash \tau \Rightarrow e}\\[1em]
+\dfrac{}{\Gamma,x:\tau \vdash \tau \Rightarrow x}~\ident{(syn-var)} &\quad&
+\dfrac{}{\Gamma \vdash \tau \Rightarrow ?_\tau}~\ident{(syn-hole)}\\[2em]
+\dfrac{\Gamma,x:\tau_1\vdash \tau_2 \Rightarrow e}{\Gamma \vdash \tau_1 \rightarrow \tau_2 \Rightarrow \lambda x.e}~\ident{(syn-lambda)}&&
+\dfrac{\Gamma \vdash \tau_1\rightarrow\tau_2 \Rightarrow e_1 \quad \Gamma \vdash \tau_1 \Rightarrow e_2}{\Gamma \vdash \tau_2 \Rightarrow e_1~e_2}~\ident{(syn-app)}\\[2em]
+\dfrac{b\in \{\ident{true},\ident{false}\} }{\Gamma \vdash \ident{bool} \Rightarrow b}~\ident{(syn-bool)} &\quad&
+\dfrac{\Gamma \vdash \tau \Rightarrow e_1 \quad \Gamma \vdash \tau \Rightarrow e_2 \quad \Gamma \vdash \ident{bool} \Rightarrow e}{\Gamma \vdash \tau \Rightarrow \ident{if}~e~\ident{then}~e_1~\ident{else}~e_2}~\ident{(syn-cond)}\\
+\end{array}
+\]
+\caption{Illustrative set of simple type-directed program synthesis rules}
+\label{fig:synths}
+\end{figure}
+
+\subparagraph{Type-directed synthesis.}
+Thanks to the equivalence between programs and proofs, techniques akin to tactic-based proof
+construction have also emerged in work on type-directed program synthesis \cite{knoth-2023-synthesis}.
+As illustrated in Figure~\ref{fig:synths}, the synthesis process can be described as a set of
+rules of the form $\Gamma \vdash \tau \Rightarrow e$ that describe ways of synthesizing
+expressions $e$ of a type $\tau$. Existing implementations of the mechanism typically aim to
+automate program synthesis and use more precise type information, such as refinement types
+\cite{polikarpova-2016-synthesis} and graded types \cite{hughes-2024-synthesis}, or include
+examples \cite{osera-2015-synthesis}. However, the same rules could be used to guide an
+interactive choose-your-own-adventure system. If the interaction was invoked to fill a typed
+hole $?_\tau$ in a context $\Gamma$, the system could collect multiple $e$ such that
+$\Gamma \vdash \tau \Rightarrow e$ and offer a choice of such options. Note that the definition
+in Figure~\ref{fig:synths} synthesizes sub-expressions recursively, but a choose-your-own-adventure
+system may always fill those with a typed hole using (\ident{syn-hole}).
+
+\subparagraph{Formal model.}
+From the perspective of user interaction, a proof assistant where a user interactively constructs
+a term of a given type is very similar to an interactive tool for type-directed program synthesis.
+The key difference being that theorem provers like Idris and Agda use rich dependent type theories.
+
+For example, consider a system akin to Idris where the user aims to construct a term $e$ of type
+$\tau$. The term may contain typed holes written as $?_\tau$ and a relation $\Gamma\vdash \tau\Rightarrow e$
+provides ways of synthesizing terms of type $\tau$. We again write $E[?_\tau]$ for a
+completion context containing a (typed) hole; we assume that the variables $\Gamma$ available in
+the completion context of the hole can be obtained using $\ident{vars}(E[\_])$.
+
+To model a choose-your-own-adventure interaction akin to Idris, the state of the system
+would be the term $e$ itself, initially a typed hole. The $\choices$ operation synthesizes
+possible completions using $\Rightarrow$ (restricted, e.g., to only generate terms of a certain
+maximum size) and offers the resulting terms as possible completions. The identifiers $\iota$
+could be based either on the tactic name (rule name) or show a preview of the resulting term.
+The $\choices$ operation suggests ways to fill a hole in the term:
+\[
+\begin{array}{rcl}
+\choices(E[?_\tau]) &=& \{\; \iota \mapsto e ~|~ \forall e\;.\; \ident{vars}(E[\_]) \vdash \tau \Rightarrow e_i \}\\
+\select(e) &=& e
+\end{array}
+\]
+The $\choices$ operation synthesizes possible terms of a type required by the hole. Since the
+state $\sigma$ is a term, the $\select$ operation simply returns it. The definition models the
+external embedding of the interaction, i.e., a system that behaves according to Idris. It
+constructs the term, but does not record the completion choices. That said, a system akin to Coq
+that constructs a sequence of tactics could be modelled too if the state was a sequence of
+tactics and $\choices$ appended the available tactics as options to the end of the current sequence.
 
@@
 
+ways of integrating automation with manual interaction
+- confirm each step vs. auto and retrace
+
 \cite{rein-2019-exploratory}
+
+\cite{verter-2024-mixed}
 

### 2025-02-20 paper.tex
Properties part 1

@@
 \subsection{Data exploration}
+\label{sec:examples-data}
 
@@
 \subsection{Type providers}
+\label{sec:examples-tps}
+
 The type provider mechanism in F\# operates at the level of the type system. It is not a merely an
@@
 \[
-\select_\mathbb{C}(e . \iota_1 . \lbrack\ldots\rbrack . \iota_n, C) = e . \iota_1 . \ldots . \iota_n
+\select_\mathbb{C}(e . \iota_1 . \lbrack\ldots\rbrack . \iota_n, C) = e . \iota_1 . \lbrack\ldots\rbrack . \iota_n
 \]
@@
 \subsection{AI assistants}
+\label{sec:examples-aias}
+
 AI assistants guide the analyst through a data wrangling task. They
@@
 \subsection{Theorem proving}
+\label{sec:examples-thm}
 
@@
 
-~
+The choose-your-own-adventure calculus lets us precisely discuss differences between how different
+programming systems interact with the user. We saw this in Section~\ref{sec:calculus}, which
+defines internal and external mode of embedding to distinguish between systems where the
+interaction leaves a reconstructible trace in the constructed program and systems where it does not.
+In this section, we make precise two properties that were introduced informally in the context of
+data exploration in The Gamma \cite{petricek-2022-thegamma}.
+
+The choose-your-own-adventure system for data exploration in The Gamma is \emph{correct}, which means that
+all programs that a user can construct using the system, by repeatedly choosing from the
+auto-completion list, are well-typed. The system is also \emph{complete}, meaning that the user can
+use auto-completion to construct all possible programs. That is, there are no well-typed programs
+that cannot be constructed interactively, by repeatedly choosing options from the offered list of
+choices.
+
+\subparagraph{Correctness.}
+The notions of correctness and completeness can, in general, be defined for any
+choose-your-own-adventure systems with respect to some system-specific distinction between
+correct and incorrect expressions. We write $\mathcal{E} \subseteq E$ for the subset
+of correct expressions.
+
+For some systems, the set of correct expressions $\mathcal{E}$
+is a set of all well-typed expressions. For some systems, we may additionally want the set
+of correct expressions $\mathcal{E}$ to be hole-free, i.e., only programs that can run (or
+complete proofs) are correct. For systems where the completion is string-based, we may treat
+all syntactically-correct programs as correct.
+
+\begin{definition}[Correctness]
+Assume that $\mathcal{E}\subseteq E$ is a subset of correct expressions,
+a choose-your-own-adventure system is correct with respect to $\mathcal{E}$ if:
+\begin{itemize}
+\item $\forall \sigma_1,..,\sigma_n$ and $\iota_i,..,\iota_n$ such that
+  $\iota_i\mapsto\sigma_i \in \choices(\sigma_{i-1})$ it is the case
+that $\select(\sigma_i) \in \mathcal{E}$.
+\end{itemize}
+\end{definition}
+
+The definition states that, if we make any sequence of choices that start from an initial state
+$\sigma_0$ and result in intermediate states $\sigma_1, \sigma_2, \ldots$, then the programs
+we could generate from any of the intermediate states are correct.
+
+The property depends on what we choose as the subset of correct expressions
+$\mathcal{E}$. Trivially, all systems are correct with respect to $\mathcal{E}=E$.
+However, the systems discussed in Section~\ref{sec:examples} are all correct with respect
+to non-trivial choices:
+
+\begin{itemize}
+\setlength{\itemsep}{5pt}
+\item For the data exploration system discussed in Section~\ref{sec:examples-data}, we say that
+  correct expressions are those where the parameters of all operations are fully-specified.
+  That is, no operation requires further arguments. With respect to this definition,
+  the system is correct. However, this is the case because the $\select$ operation drops the last
+  operation if it is not fully-specified. If $\select$ returned all operations, including the
+  partially constructed (but not yet completed) one, the system would not be correct.
+
+\item For type providers (Section~\ref{sec:examples-tps}), correct expressions are those that
+  are well-typed. With respect to this definition, the system is correct because the $\choices$
+  operation offers available members based on the type information. This also holds for the
+  type provider behind The Gamma. In The Gamma, the generated members collect operation parameters
+  and only invoke the operation once all parameters are known.
+
+\item In the case of AI assistants (Section~\ref{sec:examples-aias}) the correctness of the system
+  depends on the expressions returned by the optimization algorithm ($\argmax$) from the
+  set of all possible cleaning scripts $E_{\boldsymbol{c}}$. In general, the algorithm can return
+  any $e\in E_{\boldsymbol{c}}$ and so system correctness is a matter of definition. The system
+  is correct if and only if $E_{\boldsymbol{c}}\subseteq\mathcal{E}$ for all possible sets of
+  constraints $\boldsymbol{c}$. In practice, it is more important that the constraints generated
+  by $\choices$ are well-formed.
+
+\item For the interactive system based on type-directed synthesis (Section~\ref{sec:examples-thm}),
+  correct expressions are those that are well-typed. The system is correct if the synthesis rules
+  are sound~\cite{osera-2015-synthesis}, that is if $\Gamma \vdash \tau \Rightarrow e$ then also
+  $\Gamma \vdash e : \tau$. Note that a correct choose-your-own-adventure system can be defined
+  even using unsound synthesis rules -- it would be sufficient to filter the recommended
+  expressions in $\choices$ to the ones that are well-typed.
+\end{itemize}
+
+There may be useeful systems that violate the correctness property. An tool based on a large
+language model (LLM) may, for example, generate code with errors that the programmer can then
+correct. A more interesting case are cases like the data exploration systems discussed above
+where the program only becomes correct after multiple subsequent choices are made, for example
+to fully specify arguments of an operation.
+
+\subparagraph{Eventual correctness.}
+The data exploration system discussed in Section~\ref{sec:examples-data} ensures correctness
+by dropping the last, not fully-specified, operation in the $\select$ operation. As a result,
+it does not support internal embedding. If the operation is dropped, we cannot reconstruct it
+from the generated source code. Generating code that includes the not-fully-specified operation
+allows external embedding, but makes the system incorrect. It would still satisfy a weaker
+definition of (eventual) correctness:
+
+\begin{definition}[Eventual correctness]\label{def:eventual}
+Assume that $\mathcal{E}\subseteq E$ is a subset of correct expressions,
+a choose-your-own-adventure system is eventually correct with respect to $\mathcal{E}$ if:
+\begin{itemize}
+\item For any sequence $\sigma_1,\ldots,\sigma_k$ and $\iota_1,\ldots,\iota_k$ such that
+  $\forall i\!\in\! 1\ldots k\;.\; \iota_i\mapsto\sigma_i \in \choices(\sigma_{i-1})$
+  there exists an extension $\sigma_{k+1},\ldots,\sigma_n$ and $\iota_{k+1},\ldots,\iota_n$ such that
+  $\select(\sigma_n) \in \mathcal{E}$ and
+  $\forall i\!\in\! k\!+\!1\ldots n\;.\;\iota_i\mapsto\sigma_i \in \choices(\sigma_{i-1})$.
+\end{itemize}
+\end{definition}
+
+Eventual correctness models systems where some sequences of choices result in invalid programs,
+but it is always possible to make further choices to reach a valid program. In general, it is
+always possible to turn an eventually correct system into a correct one:
+
+\begin{enumerate}
+\setlength{\itemsep}{5pt}
+\item As in the case of the data exploration, the system can remember the last state for which
+  the $\select$ operation returned a correct program and use it until the next correct state is
+  reached. This makes any eventually correct choose-your-own-adventure system correct, but it
+  does not support internal embedding.
+
+\item Alternatively, we can construct a system that collapses all sequence of temporarilly invalid states
+  $\sigma_1, \ldots, \sigma_n$ identified by $\iota_1, \ldots, \sigma_n$ where
+  $\forall i\!\in\! 1\ldots n\!-\! 1\;.\;\select(\sigma_i)\notin\mathcal{E}$ and $\select(\sigma_n)\in\mathcal{E}$
+  into a single option $\iota_1\ldots\iota_n \mapsto \sigma_n$  designated by a joined identifier.
+  This makes the system correct and also preserves external embedding, but it potentially
+  generates too many choices that are difficult to understand.
+\end{enumerate}
+
+There is more to be said about correctness of interactive programming systems, but the
+conceptual framework provided by the choose-your-own-adventure calculus makes it possible to
+take the first step. Similarly, the model lets us formally define the second property
+mentioned earlier.
+
+\subparagraph{Completeness.}
+
+todo
+
+..
+
+..
+
+..
+
+\begin{definition}[Completeness]
+Assume that $E$ is a set of all possible expressions in a language
+and $\mathcal{E}\subseteq E$ is a set of all expressions that are correct with respect to some
+case-specific notion of correctness (e.g. well-typed). A choose-your-own-adventure system is
+complete if:
+
+$\forall e\in \mathcal{E}.~\exists \sigma_1,\ldots,\sigma_n$ such that
+$\sigma_i \in \choices(\sigma_{i-1})$ and $e=\select(\sigma_n)$.
+\end{definition}
+
+That is, for any correct program, there is a sequence of choices that leads to a system state
+that $\select$ turns into the given program. This is a more subtle property that not all of my
+above examples have:
+
+* In The Gamma, the programs that can be generated are restricted - you can only use a fixed set
+  of aggregation operations (rather than writing your own) and only a restricted set of parameters
+  (sorting by a key, but not based on a custom expression). For those, the type provider is
+  complete. However, if we treated a more general-purpose query language as the underlying
+  language, the provider would not be complete.
+
+* In AI assistants, the system offers a set of constraints that is generated based on the selected
+  expression, but this does not let you construct arbitrary constraints. Moreover, because the
+  $\select$ operation is AI-based, it is not guaranteed that there is a way to get it to generate
+  a specific program (unless we can supply constraints that restrict the set of programs $E_\boldsymbol{c}$
+  to just a single program).
+
+* In interactive theorem prover, we could possibly offer all possible ways of filling a hole
+  (up to renaming), but this would not be very practical. It is more likely that
+  tactics will only generate a subset of valid proof/program steps and the user has to write some
+  other steps manually.
+
+The nice thing about the choose-your-own-adventure formalism is that it also lets us talk about
+(and think about!) properties that are more specifically about the user interaction with an
+interactive programming system.
 
@@
 
-\newpage
-
-\section{Typesetting instructions -- Summary}
-\label{sec:typesetting-summary}
-
-LIPIcs is a series of open access high-quality conference proceedings across all fields in informatics established in cooperation with Schloss Dagstuhl.
-In order to do justice to the high scientific quality of the conferences that publish their proceedings in the LIPIcs series, which is ensured by the thorough review process of the respective events, we believe that LIPIcs proceedings must have an attractive and consistent layout matching the standard of the series.
-Moreover, the quality of the metadata, the typesetting and the layout must also meet the requirements of other external parties such as indexing service, DOI registry, funding agencies, among others.
-The guidelines contained in this document serve as the baseline for the authors, editors, and the publisher to create documents that meet as many different requirements as possible.
-
-Please comply with the following instructions when preparing your article for a LIPIcs proceedings volume.
-\paragraph*{Minimum requirements}
-
-\begin{itemize}
-\item Use pdflatex and an up-to-date \LaTeX{} system.
-\item Use further \LaTeX{} packages and custom made macros carefully and only if required.
-\item Use the provided sectioning macros: \verb+\section+, \verb+\subsection+, \verb+\subsubsection+, \linebreak \verb+\paragraph+, \verb+\paragraph*+, and \verb+\subparagraph*+.
-\item Provide suitable graphics of at least 300dpi (preferably in PDF format).
-\item Use BibTeX and keep the standard style (\verb+plainurl+) for the bibliography.
-\item Please try to keep the warnings log as small as possible. Avoid overfull \verb+\hboxes+ and any kind of warnings/errors with the referenced BibTeX entries.
-\item Use a spellchecker to correct typos.
-\end{itemize}
-
-\paragraph*{Mandatory metadata macros}
-Please set the values of the metadata macros carefully since the information parsed from these macros will be passed to publication servers, catalogues and search engines.
-Avoid placing macros inside the metadata macros. The following metadata macros/environments are mandatory:
-\begin{itemize}
-\item \verb+\title+ and, in case of long titles, \verb+\titlerunning+.
-\item \verb+\author+, one for each author, even if two or more authors have the same affiliation.
-\item \verb+\authorrunning+ and \verb+\Copyright+ (concatenated author names)\\
-The \verb+\author+ macros and the \verb+\Copyright+ macro should contain full author names (especially with regard to the first name), while \verb+\authorrunning+ should contain abbreviated first names.
-\item \verb+\ccsdesc+ (ACM classification, see \url{https://www.acm.org/publications/class-2012}).
-\item \verb+\keywords+ (a comma-separated list of keywords).
-\item \verb+\relatedversion+ (if there is a related version, typically the ``full version''); please make sure to provide a persistent URL, e.\,g., at arXiv.
-\item \verb+\begin{abstract}...\end{abstract}+ .
-\end{itemize}
-
-\paragraph*{Please do not \ldots} %Do not override the \texttt{\seriesstyle}-defaults}
-Generally speaking, please do not override the \texttt{lipics-v2021}-style defaults.
-To be more specific, a short checklist also used by Dagstuhl Publishing during the final typesetting is given below.
-In case of \textbf{non-compliance} with these rules Dagstuhl Publishing will remove the corresponding parts of \LaTeX{} code and \textbf{replace it with the \texttt{lipics-v2021} defaults}.
-In serious cases, we may reject the LaTeX-source and expect the corresponding author to revise the relevant parts.
-\begin{itemize}
-\item Do not use a different main font. (For example, the \texttt{times} package is forbidden.)
-\item Do not alter the spacing of the \texttt{lipics-v2021.cls} style file.
-\item Do not use \verb+enumitem+ and \verb+paralist+. (The \texttt{enumerate} package is preloaded, so you can use
- \verb+\begin{enumerate}[(a)]+ or the like.)
-\item Do not use ``self-made'' sectioning commands (e.\,g., \verb+\noindent{\bf My+ \verb+Paragraph}+).
-\item Do not hide large text blocks using comments or \verb+\iffalse+ $\ldots$ \verb+\fi+ constructions.
-\item Do not use conditional structures to include/exclude content. Instead, please provide only the content that should be published -- in one file -- and nothing else.
-\item Do not wrap figures and tables with text. In particular, the package \texttt{wrapfig} is not supported.
-\item Do not change the bibliography style. In particular, do not use author-year citations. (The
-\texttt{natbib} package is not supported.)
-\end{itemize}
-
-\enlargethispage{\baselineskip}
-
-This is only a summary containing the most relevant details.
-Please read the complete document ``LIPIcs: Instructions for Authors and the \texttt{lipics-v2021} Class'' for all details and don't hesitate to contact Dagstuhl Publishing (\url{mailto:publishing@dagstuhl.de}) in case of questions or comments:
-\href{http://drops.dagstuhl.de/styles/lipics-v2021/lipics-v2021-authors/lipics-v2021-authors-guidelines.pdf}{\texttt{http://drops.dagstuhl.de/styles/lipics-v2021/\newline lipics-v2021-authors/lipics-v2021-authors-guidelines.pdf}}
-
-\section{Lorem ipsum dolor sit amet}
-
-Lorem ipsum dolor sit amet, consectetur adipiscing elit \cite{DBLP:journals/cacm/Knuth74}.
-Praesent convallis orci arcu, eu mollis dolor.
-Aliquam eleifend suscipit lacinia.
-Maecenas quam mi, porta ut lacinia sed, convallis ac dui.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Suspendisse potenti.
-Donec eget odio et magna ullamcorper vehicula ut vitae libero.
-Maecenas lectus nulla, auctor nec varius ac, ultricies et turpis.
-Pellentesque id ante erat.
-In hac habitasse platea dictumst.
-Curabitur a scelerisque odio.
-Pellentesque elit risus, posuere quis elementum at, pellentesque ut diam.
-Quisque aliquam libero id mi imperdiet quis convallis turpis eleifend.
-
-\begin{lemma}[Lorem ipsum]
-\label{lemma:lorem}
-Vestibulum sodales dolor et dui cursus iaculis.
-Nullam ullamcorper purus vel turpis lobortis eu tempus lorem semper.
-Proin facilisis gravida rutrum.
-Etiam sed sollicitudin lorem.
-Proin pellentesque risus at elit hendrerit pharetra.
-Integer at turpis varius libero rhoncus fermentum vitae vitae metus.
-\end{lemma}
-
-\begin{proof}
-Cras purus lorem, pulvinar et fermentum sagittis, suscipit quis magna.
-
-\proofsubparagraph*{Just some paragraph within the proof.}
-Nam liber tempor cum soluta nobis eleifend option congue nihil imperdiet doming id quod mazim placerat facer possim assum.
-Lorem ipsum dolor sit amet, consectetuer adipiscing elit, sed diam nonummy nibh euismod tincidunt ut laoreet dolore magna aliquam erat volutpat.
-\begin{claim}
-content...
-\end{claim}
-\begin{claimproof}
-content...
-    \begin{enumerate}
-        \item abc abc abc \claimqedhere{}
-    \end{enumerate}
-\end{claimproof}
-
-\end{proof}
-
-\begin{corollary}[Curabitur pulvinar, \cite{DBLP:books/mk/GrayR93}]
-\label{lemma:curabitur}
-Nam liber tempor cum soluta nobis eleifend option congue nihil imperdiet doming id quod mazim placerat facer possim assum.
-Lorem ipsum dolor sit amet, consectetuer adipiscing elit, sed diam nonummy nibh euismod tincidunt ut laoreet dolore magna aliquam erat volutpat.
-\end{corollary}
-
-\begin{proposition}\label{prop1}
-This is a proposition
-\end{proposition}
-
-\autoref{prop1} and \cref{prop1} \ldots
-
-\subsection{Curabitur dictum felis id sapien}
-
-Curabitur dictum \cref{lemma:curabitur} felis id sapien \autoref{lemma:curabitur} mollis ut venenatis tortor feugiat.
-Curabitur sed velit diam.
-Integer aliquam, nunc ac egestas lacinia, nibh est vehicula nibh, ac auctor velit tellus non arcu.
-Vestibulum lacinia ipsum vitae nisi ultrices eget gravida turpis laoreet.
-Duis rutrum dapibus ornare.
-Nulla vehicula vulputate iaculis.
-Proin a consequat neque.
-Donec ut rutrum urna.
-Morbi scelerisque turpis sed elit sagittis eu scelerisque quam condimentum.
-Pellentesque habitant morbi tristique senectus et netus et malesuada fames ac turpis egestas.
-Aenean nec faucibus leo.
-Cras ut nisl odio, non tincidunt lorem.
-Integer purus ligula, venenatis et convallis lacinia, scelerisque at erat.
-Fusce risus libero, convallis at fermentum in, dignissim sed sem.
-Ut dapibus orci vitae nisl viverra nec adipiscing tortor condimentum \cite{DBLP:journals/cacm/Dijkstra68a}.
-Donec non suscipit lorem.
-Nam sit amet enim vitae nisl accumsan pretium.
-
-\begin{lstlisting}[caption={Useless code.},label=list:8-6,captionpos=t,float,abovecaptionskip=-\medskipamount]
-for i:=maxint to 0 do
-begin
-    j:=square(root(i));
-end;
-\end{lstlisting}
-
-\subsection{Proin ac fermentum augue}
-
-Proin ac fermentum augue.
-Nullam bibendum enim sollicitudin tellus egestas lacinia euismod orci mollis.
-Nulla facilisi.
-Vivamus volutpat venenatis sapien, vitae feugiat arcu fringilla ac.
-Mauris sapien tortor, sagittis eget auctor at, vulputate pharetra magna.
-Sed congue, dui nec vulputate convallis, sem nunc adipiscing dui, vel venenatis mauris sem in dui.
-Praesent a pretium quam.
-Mauris non mauris sit amet eros rutrum aliquam id ut sapien.
-Nulla aliquet fringilla sagittis.
-Pellentesque eu metus posuere nunc tincidunt dignissim in tempor dolor.
-Nulla cursus aliquet enim.
-Cras sapien risus, accumsan eu cursus ut, commodo vel velit.
-Praesent aliquet consectetur ligula, vitae iaculis ligula interdum vel.
-Integer faucibus faucibus felis.
-
-\begin{itemize}
-\item Ut vitae diam augue.
-\item Integer lacus ante, pellentesque sed sollicitudin et, pulvinar adipiscing sem.
-\item Maecenas facilisis, leo quis tincidunt egestas, magna ipsum condimentum orci, vitae facilisis nibh turpis et elit.
-\end{itemize}
-
-\begin{remark}
-content...
-\end{remark}
-
-\section{Pellentesque quis tortor}
-
-Nec urna malesuada sollicitudin.
-Nulla facilisi.
-Vivamus aliquam tempus ligula eget ornare.
-Praesent eget magna ut turpis mattis cursus.
-Aliquam vel condimentum orci.
-Nunc congue, libero in gravida convallis \cite{DBLP:conf/focs/HopcroftPV75}, orci nibh sodales quam, id egestas felis mi nec nisi.
-Suspendisse tincidunt, est ac vestibulum posuere, justo odio bibendum urna, rutrum bibendum dolor sem nec tellus.
-
-\begin{lemma} [Quisque blandit tempus nunc]
-Sed interdum nisl pretium non.
-Mauris sodales consequat risus vel consectetur.
-Aliquam erat volutpat.
-Nunc sed sapien ligula.
-Proin faucibus sapien luctus nisl feugiat convallis faucibus elit cursus.
-Nunc vestibulum nunc ac massa pretium pharetra.
-Nulla facilisis turpis id augue venenatis blandit.
-Cum sociis natoque penatibus et magnis dis parturient montes, nascetur ridiculus mus.
-\end{lemma}
-
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-
-\section{Morbi eros magna}
-
-Morbi eros magna, vestibulum non posuere non, porta eu quam.
-Maecenas vitae orci risus, eget imperdiet mauris.
-Donec massa mauris, pellentesque vel lobortis eu, molestie ac turpis.
-Sed condimentum convallis dolor, a dignissim est ultrices eu.
-Donec consectetur volutpat eros, et ornare dui ultricies id.
-Vivamus eu augue eget dolor euismod ultrices et sit amet nisi.
-Vivamus malesuada leo ac leo ullamcorper tempor.
-Donec justo mi, tempor vitae aliquet non, faucibus eu lacus.
-Donec dictum gravida neque, non porta turpis imperdiet eget.
-Curabitur quis euismod ligula.
-
-%%
-%% Bibliography
-%%
-
-%% Please use bibtex,
-
-\bibliography{lipics-v2021-sample-article}
-
-\appendix
-
-\section{Styles of lists, enumerations, and descriptions}\label{sec:itemStyles}
-
-List of different predefined enumeration styles:
-
-\begin{itemize}
-\item \verb|\begin{itemize}...\end{itemize}|
-\item \dots
-\item \dots
-%\item \dots
-\end{itemize}
-
-\begin{enumerate}
-\item \verb|\begin{enumerate}...\end{enumerate}|
-\item \dots
-\item \dots
-%\item \dots
-\end{enumerate}
-
-\begin{alphaenumerate}
-\item \verb|\begin{alphaenumerate}...\end{alphaenumerate}|
-\item \dots
-\item \dots
-%\item \dots
-\end{alphaenumerate}
-
-\begin{romanenumerate}
-\item \verb|\begin{romanenumerate}...\end{romanenumerate}|
-\item \dots
-\item \dots
-%\item \dots
-\end{romanenumerate}
-
-\begin{bracketenumerate}
-\item \verb|\begin{bracketenumerate}...\end{bracketenumerate}|
-\item \dots
-\item \dots
-%\item \dots
-\end{bracketenumerate}
-
-\begin{description}
-\item[Description 1] \verb|\begin{description} \item[Description 1]  ...\end{description}|
-\item[Description 2] Fusce eu leo nisi.
-Cras eget orci neque, eleifend dapibus felis.
-Duis et leo dui.
-Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\item[Description 3]  \dots
-%\item \dots
-\end{description}
-
-\cref{testenv-proposition} and \autoref{testenv-proposition} ...
-
-\section{Theorem-like environments}\label{sec:theorem-environments}
-
-List of different predefined enumeration styles:
-
-\begin{theorem}\label{testenv-theorem}
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\end{theorem}
-
-\begin{lemma}\label{testenv-lemma}
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\end{lemma}
-
-\begin{corollary}\label{testenv-corollary}
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\end{corollary}
-
-\begin{proposition}\label{testenv-proposition}
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\end{proposition}
-
-\begin{conjecture}\label{testenv-conjecture}
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\end{conjecture}
-
-\begin{observation}\label{testenv-observation}
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\end{observation}
-
-\begin{exercise}\label{testenv-exercise}
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\end{exercise}
-
-\begin{definition}\label{testenv-definition}
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\end{definition}
-
-\begin{example}\label{testenv-example}
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\end{example}
-
-\begin{note}\label{testenv-note}
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\end{note}
-
-\begin{note*}
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\end{note*}
-
-\begin{remark}\label{testenv-remark}
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\end{remark}
-
-\begin{remark*}
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\end{remark*}
-
-\begin{claim}\label{testenv-claim}
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\end{claim}
-
-\begin{claim*}\label{testenv-claim2}
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\end{claim*}
-
-\begin{proof}
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\end{proof}
-
-\begin{claimproof}
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\end{claimproof}
-
 \end{document}

### 2025-02-21 paper.tex
Mixed-initiative systems

@@
 \newcommand{\ident}[1]{\textsf{#1}}
+\newcommand{\kvd}[1]{\textsf{\bfseries #1}}
 \newcommand{\ddident}[1]{\guillemotleft\ident{#1}\guillemotright}
@@
 \newcommand{\select}{\textnormal{\ident{choose}}}
+\newcommand{\suggest}{\textnormal{\ident{suggest}}}
 \newcommand{\choices}{\textnormal{\ident{choices}}}
@@
 
-todo
+TODO: Say somewhere that this also formalizes things like mixed-initiative interaction and
+programming by demonstration
 
@@
 
-\vspace{-0.25em}
+\vspace{-0.5em}
 \raggedright
@@
 \item $\ident{init}(E[?]) = \sigmaN$ obtains the initial state of a choose-your-own-interaction system,
-\item $\sigma_n$ is a system state such that $\forall i\in{1\ldots n}.(\iota_i\mapsto\sigma_i)\in\choices{(\sigma_{n-1})}$,
+\item $\sigma_n$ is a system state such that $\forall i\!\in\! 1\ldots n\;.\;(\iota_i\mapsto\sigma_i)\in\choices{(\sigma_{n-1})}$,
   i.e.,~the user\\ makes a series of choices resulting in a final state of the system $\sigma_n$,
@@
 
-The choose-your-own-adventure calculus lets us precisely discuss differences between how different
+The choose-your-own-adventure calculus lets us precisely compare how how different
 programming systems interact with the user. We saw this in Section~\ref{sec:calculus}, which
@@
 
-The choose-your-own-adventure system for data exploration in The Gamma is \emph{correct}, which means that
+The choose-your-own-adventure system for data exploration in The Gamma is \emph{correct}, meaning that
 all programs that a user can construct using the system, by repeatedly choosing from the
 auto-completion list, are well-typed. The system is also \emph{complete}, meaning that the user can
-use auto-completion to construct all possible programs. That is, there are no well-typed programs
+use auto-completion to construct all possible programs. In other words, there are no well-typed programs
 that cannot be constructed interactively, by repeatedly choosing options from the offered list of
@@
 \subparagraph{Correctness.}
-The notions of correctness and completeness can, in general, be defined for any
+The notions of correctness and completeness can be defined for any
 choose-your-own-adventure systems with respect to some system-specific distinction between
@@
 
-For some systems, the set of correct expressions $\mathcal{E}$
+For systems based on statically-typed programming languages, a reasonable choice of $\mathcal{E}$
 is a set of all well-typed expressions. For some systems, we may additionally want the set
 of correct expressions $\mathcal{E}$ to be hole-free, i.e., only programs that can run (or
-complete proofs) are correct. For systems where the completion is string-based, we may treat
-all syntactically-correct programs as correct.
+represent complete proofs) are correct. For systems where the completion is string-based,
+we may treat all syntactically-correct programs as correct.
 
 \begin{definition}[Correctness]
-Assume that $\mathcal{E}\subseteq E$ is a subset of correct expressions,
-a choose-your-own-adventure system is correct with respect to $\mathcal{E}$ if:
+Assume that $\mathcal{E}\subseteq E$ is a subset of correct expressions.
+A~choose-your-own-adventure system is correct with respect to $\mathcal{E}$ if and only if:
 \begin{itemize}
@@
 The definition states that, if we make any sequence of choices that start from an initial state
-$\sigma_0$ and result in intermediate states $\sigma_1, \sigma_2, \ldots$, then the programs
+$\sigma_0$ and result in intermediate states $\sigma_1, \ldots, \sigma_n$ then the programs
 we could generate from any of the intermediate states are correct.
@@
 $\mathcal{E}$. Trivially, all systems are correct with respect to $\mathcal{E}=E$.
-However, the systems discussed in Section~\ref{sec:examples} are all correct with respect
-to non-trivial choices:
+However, the systems discussed in Section~\ref{sec:examples} are, with one exception,
+correct with respect to a non-trivial set of correct expressions:
 
@@
   That is, no operation requires further arguments. With respect to this definition,
-  the system is correct. However, this is the case because the $\select$ operation drops the last
+  the system is correct. However, this is the case only because the $\select$ operation drops the last
   operation if it is not fully-specified. If $\select$ returned all operations, including the
-  partially constructed (but not yet completed) one, the system would not be correct.
+  partially constructed (but not yet completed) operation, the system would not be correct.
 
@@
   are well-typed. With respect to this definition, the system is correct because the $\choices$
-  operation offers available members based on the type information. This also holds for the
+  operation offers available members based on the type information. This reasoning also applies to the
   type provider behind The Gamma. In The Gamma, the generated members collect operation parameters
@@
 \item In the case of AI assistants (Section~\ref{sec:examples-aias}) the correctness of the system
-  depends on the expressions returned by the optimization algorithm ($\argmax$) from the
+  depends on the expressions returned by the optimization algorithm ($\mathit{arg\,max}$) from the
   set of all possible cleaning scripts $E_{\boldsymbol{c}}$. In general, the algorithm can return
@@
 There may be useeful systems that violate the correctness property. An tool based on a large
-language model (LLM) may, for example, generate code with errors that the programmer can then
-correct. A more interesting case are cases like the data exploration systems discussed above
-where the program only becomes correct after multiple subsequent choices are made, for example
+language model (LLM) may generate code with errors that the programmer can later
+correct. A more interesting case would be a data exploration system, like the one discussed above,
+where programs only become correct after multiple subsequent choices are made, for example
 to fully specify arguments of an operation.
@@
 The data exploration system discussed in Section~\ref{sec:examples-data} ensures correctness
-by dropping the last, not fully-specified, operation in the $\select$ operation. As a result,
-it does not support internal embedding. If the operation is dropped, we cannot reconstruct it
-from the generated source code. Generating code that includes the not-fully-specified operation
-allows external embedding, but makes the system incorrect. It would still satisfy a weaker
-definition of (eventual) correctness:
+by dropping the last non-fully-specified operation in the $\select$ operation. As a result,
+it is correct, but it does not support internal embedding. If the operation is dropped, we cannot
+reconstruct the full sequence of interactions from the generated source code. Generating code
+that includes the non-fully-specified operation allows external embedding, but makes the system
+incorrect. It would still satisfy a weaker definition of (eventual) correctness.
 
@@
 Eventual correctness models systems where some sequences of choices result in invalid programs,
-but it is always possible to make further choices to reach a valid program. In general, it is
+but it is always possible to make further choices to reach a valid program. It is
 always possible to turn an eventually correct system into a correct one:
@@
 \item As in the case of the data exploration, the system can remember the last state for which
-  the $\select$ operation returned a correct program and use it until the next correct state is
+  the $\select$ operation returned a correct program and use it in $\select$ until the next correct state is
   reached. This makes any eventually correct choose-your-own-adventure system correct, but it
@@
   $\forall i\!\in\! 1\ldots n\!-\! 1\;.\;\select(\sigma_i)\notin\mathcal{E}$ and $\select(\sigma_n)\in\mathcal{E}$
-  into a single option $\iota_1\ldots\iota_n \mapsto \sigma_n$  designated by a joined identifier.
+  into a single option $\iota' \mapsto \sigma_n$ where $\iota'$ is produced by concatenating
+  identifiers $\iota_1,\ldots,\iota_n$.
   This makes the system correct and also preserves external embedding, but it potentially
-  generates too many choices that are difficult to understand.
+  generates too many choices that are difficult to navigate.
 \end{enumerate}
@@
 conceptual framework provided by the choose-your-own-adventure calculus makes it possible to
-take the first step. Similarly, the model lets us formally define the second property
-mentioned earlier.
+take the first step. Similarly, the model lets us formally define completeness.
 
 \subparagraph{Completeness.}
-
-todo
-
-..
-
-..
-
-..
+The programming language used in The Gamma allows users to write scripts that also use let bindings
+and method calls. However, for chains of member accesses which the user can construct using
+a type provider, it is possible to construct any chain just by repeatedly choosing options from
+the offered list. This is captured as the completeness property of a choose-your-own-adventure system.
 
 \begin{definition}[Completeness]
-Assume that $E$ is a set of all possible expressions in a language
-and $\mathcal{E}\subseteq E$ is a set of all expressions that are correct with respect to some
-case-specific notion of correctness (e.g. well-typed). A choose-your-own-adventure system is
-complete if:
-
-$\forall e\in \mathcal{E}.~\exists \sigma_1,\ldots,\sigma_n$ such that
-$\sigma_i \in \choices(\sigma_{i-1})$ and $e=\select(\sigma_n)$.
+Assume that $\mathcal{E}\subseteq E$ is a subset of correct expressions.
+A~choose-your-own-adventure system is complete with respect to $\mathcal{E}$ if and only if:
+\begin{itemize}
+\item $\forall e\!\in\!\mathcal{E}\,.\,\exists \sigma_1,..,\sigma_n$ and $\iota_i,..,\iota_n$ such that
+  $\iota_i\mapsto\sigma_i \in \choices(\sigma_{i-1})$ and $e=\select(\sigma_n)$.
+\end{itemize}
 \end{definition}
 
-That is, for any correct program, there is a sequence of choices that leads to a system state
-that $\select$ turns into the given program. This is a more subtle property that not all of my
-above examples have:
+A system is complete if, for any correct program, there is a sequence of choices that can be
+used to construct the given program. This is a more subtle property than correctness and
+it does not hold for all the examples dicusssed in Section~\ref{sec:examples}.
 
-* In The Gamma, the programs that can be generated are restricted - you can only use a fixed set
-  of aggregation operations (rather than writing your own) and only a restricted set of parameters
-  (sorting by a key, but not based on a custom expression). For those, the type provider is
-  complete. However, if we treated a more general-purpose query language as the underlying
-  language, the provider would not be complete.
+\begin{itemize}
+\setlength{\itemsep}{5pt}
+\item The data exploration system described in Section~\ref{sec:examples-data} would be complete
+  only if the underlying query language had a fixed set of operations, a fixed set of aggregation
+  operations (rather than letting users write their own) and a fixed set of values for each
+  parameter (sorting by a key, but not based on a custom expression). A completion system for a
+  more general-purpose query language would be incomplete.
 
-* In AI assistants, the system offers a set of constraints that is generated based on the selected
-  expression, but this does not let you construct arbitrary constraints. Moreover, because the
-  $\select$ operation is AI-based, it is not guaranteed that there is a way to get it to generate
-  a specific program (unless we can supply constraints that restrict the set of programs $E_\boldsymbol{c}$
-  to just a single program).
+\item For type providers (Section~\ref{sec:examples-tps}), the completion mechanism is complete,
+  because it offers all available members of the type (and ``.'' has to be followed by a member
+  access). Consequently, the type provider in The Gamma is also complete. Even if the underlying
+  query language is more expressive, it is hidden from the user and the system offers all
+  available members.
 
-* In interactive theorem prover, we could possibly offer all possible ways of filling a hole
-  (up to renaming), but this would not be very practical. It is more likely that
-  tactics will only generate a subset of valid proof/program steps and the user has to write some
-  other steps manually.
+\item For AI assistants (Section~\ref{sec:examples-aias}), the system offers a set of constraints
+  based on the current inferred program. It does not let the user construct arbitrary constraints.
+  Moreover, because the $\mathit{arg\,max}$ operation used in $\select$ performs statistical
+  optimization, there is no guarantee that it can be used to generate a specific program.
+  The system is only complete if it is possible to choose a constraint set $\boldsymbol{c}$ that
+  restrict the set of programs $E_\boldsymbol{c}$ to a single given program. This is the case
+  only for some AI assistants (e.g., datadiff, which always offers constraints to map column to
+  any chosen other column).
 
-The nice thing about the choose-your-own-adventure formalism is that it also lets us talk about
-(and think about!) properties that are more specifically about the user interaction with an
-interactive programming system.
+\item A type-directed synthesis system (Section~\ref{sec:examples-thm}), or an interactive theorem
+  prover could offer all possible ways of filling a hole with an expression containing further
+  holes as sub-expressions. This would make the system complete (up to renaming of variables this
+  may introduce), but the great number of generated options would be impractical. A realistic
+  system would only generate a subset of most useful proof/program steps and let the user write
+  other steps manually (interactive proof construction in Idris can be seen as operating in this way).
+\end{itemize}
+
+Correctness and completeness are arguably both desirable properties of a choose-your-own-adventure
+system. Unlike for example type soundess, they are not strictly necessary in practice and are best
+seen as design trade-offs that designers should consider.
 
@@
 \label{sec:applications}
+
+The choose-your-own-adventure calculus lets us treat a wide range of interactive programming
+systems as instances of the same general pattern. This makes it possible to discuss properties
+of the systems (as done in the previous section), reuse components in system implementations,
+but also transfer ideas across different domains. In this section, we discuss three ideas that
+emerged in the context of a specific interactive system, but could be applied to any system
+based on the choose-your-own-adventure pattern.
+
+~
+
+\begin{figure}[t]
+\definecolor{shade1}{HTML}{f0f0f0}
+\colorbox{shade1}{\colorbox{white}{
+$\begin{array}{l}
+\kvd{theorem}~\ident{sum-z-rh}\!:~\kvd{forall}~n~\kvd{exists}~n+(\ident{z})=n.\\
+\kvd{theorem}~\ident{sum-s-rh}\!:~\kvd{forall}~d_1:n_1+n_2=n_3~\kvd{exists}~n_1+(\ident{s}~n_2)=(\ident{s}~n_3).\\[0.75em]
+\kvd{theorem}~\ident{sum-commutes}\!:~\kvd{forall}~d_1:n_1+n_2=n_3~\kvd{exists}~n_2+n_1=n_3\\
+\quad\;\;\; d_2: n_2 + n_1 = n_3~\kvd{by induction on}~d_1\!:\\
+\quad\colorbox{shade1}{$
+\begin{array}{l}
+  \kvd{case rule}\\[0.25em]
+  \quad\;\;\; \dfrac{}{\mathit{dzc}:~(\ident{z})+n=n}~\ident{sum-z}\\[0.25em]
+  \quad\kvd{is}\\
+  \quad\;\;\; \mathit{dz}_1: n + (\ident{z}) = n~\kvd{by theorem}~\ident{sum-z-rh}~\kvd{on}~n\\
+  \kvd{end case}\\
+  \kvd{case rule}\\[0.25em]
+  \quad\;\;\;\dfrac{\mathit{dsp}\!: n_1' + n_2 = n_3'}{\mathit{dsc}\!: (\ident{s}~n_1') + n_2 = (\ident{s}~n_3')}~\ident{sum-s}\\[0.25em]
+  \quad\kvd{is}\\
+  \quad\colorbox{white}{$\begin{array}{l}
+    \;\;\;\mathit{ds}_1\!: n_2 + n_1' = n_3'~\kvd{by induction hypothesis on}~\mathit{dsp}\\
+    \colorbox{shade1}{$\begin{array}{l}
+    \mathit{ds}_2\!: n_2 + (\ident{s} n_1') = (\ident{s} n_3')~\kvd{by theorem}~\ident{sum-s-rh}~\kvd{on}~\mathit{ds}_1\hspace{8em}~
+    \end{array}$}\\
+  \end{array}$}\\
+  \kvd{end case}\\
+\end{array}
+$}\\
+\quad\;\;\;\kvd{end induction}\\
+\kvd{end theorem}
+\end{array}
+$}}
+\caption{A proof of commutativity of $+$ in Peano arithmetic constructed using mixed-initiative
+  interaction. Parts generated by the system are highlighted with gray background.
+  After the user writes a theorem and specifies induction, the system completes the
+  first case. User then specifies how to apply the induction hypothesis and the system completes
+  the proof.}
+\label{fig:proof}
+\end{figure}
+
+\subsection{Mixed-initiative interaction}
+When using an interactive theorem prover, one typically constructs the proof manually until an
+automated strategy can fill in the remaining gaps. This is the case with Idris proof search,
+as well as Coq \texttt{auto} tactics. Richer ways of interacting exist \cite{lowe-1997-xbarnacle},
+but are less common. We developed a prototype mixed-initiative theorem prover [report citation
+omitted] that supports a more collaborative way of working where the system completes some
+steps automatically, but defers back to the user when it gets stuck.
+
+\subparagraph{Mixed-initiative theorem proving.}
+As an illustration of the mixed-initiative proving, consider Figure~\ref{fig:proof} which
+shows a proof of commutativity of $+$ in Peano arithmetic. The figure shows a version of our example
+[report citation omitted],\footnote{Adapted from \url{https://github.com/boyland/sasylf/blob/master/examples/sum.slf}}
+simplified for brevity and written using the SASyLF notation \cite{aldrich-2008-sasylf}.
+After writing proofs of \ident{sum-z-rh} and \ident{sum-s-rh} (not shown), the user states
+\ident{sum-commutes} and specifies the structure of the induction. They then invoke the automatic
+search, which completes the first case, but gets stuck in the second case, because it fails
+to apply the induction hypothesis (in our full example, the failure is more subtle). The user
+then specifies how to apply the induction hypothesis and the system automatically completes the
+proof.
+
+The interaction can be revisited from the perspective of the choose-your-own-adventure system
+discussed in Section~\ref{sec:examples-thm}. An interactive theorem prover generates possible
+completions using available tactics and offers them to the user, who chooses a tactic and
+applies it to transform the proof. In the automatic mode, the interactive theorem prover
+recursively searches through the available choices. If it finds one that results in a complete
+proof, it stops. Otherwise, it completes a number of steps (determined by some heuristic)
+and defers back to the user who chooses the next step and completes the proof manually or
+invokes the automatic search again.
+
+\subparagraph{Generalised mixed-initiative interaction.}
+The mixed-initiative mode of interaction combines manual interaction with automatic search.
+In order to support automatic search, the system needs a metric that determines whether a
+constructed program is correct (as when proving a given theorem) or whether it is an improvement
+over another equivalent program (for example when refactoring). This general way of working
+can be used in other interactive programming systems:
+
+\begin{itemize}
+\item A system for type-directed program synthesis could automatically synthesize parts of a
+  program (e.g., pattern matching and implementation for some of the cases), but ask user for
+  input in order to complete branches where solution was not found automatically.
+\item In data exploration, a system could perform automatic search through the available operations
+  in order to transform data into a more regular format, according to a suitability
+  metric as done, for example, in the Proactive Wrangling system~\cite{guo-2011-proactive}.
+\end{itemize}
+
+We can understand how mixed-initiative interactive systems as extensions
+built on top of the choose-your-own-adventure calculus. A mixed-initiative interactive system
+defines an operation \ident{suggest} that recommends a sequence of choices for a given state.
+
+\begin{definition}[Mixed-initiative system]
+A choose-your-own-adventure system supports mixed-initiative mode of interaction if it is
+equipped with an operation $\suggest$ such that:
+\begin{itemize}
+  \item $\suggest(\sigma_0)=\iota_1, \ldots, \iota_n$ such that
+    $\forall i\!\in\! 1\ldots n\;.\;\iota_i\mapsto\sigma_i\in\choices(\sigma_{i-1})$.
+\end{itemize}
+\end{definition}
+
+The definition only requires that the suggested sequence of choices is valid, but it
+does not specify how exactly the system should make the recommendations. This is specific to
+a particular system. An interactive theorem prover will try to find a proof or solve
+sub-goals, whereas data wrangling system may try to improve the structure of data.
+In practice, the suggestion should improve the quality of the constructed program so that
+$\select(\sigma_n)$ is better than $\select(\sigma_0)$, but we leave the definition flexible
+to accommodate a broader range of potential uses.
+
+\subsection{LLM assistant for data exploration}
+
+\subsection{Programming by demonstration}
 
@@
 
+% cite: https://link.springer.com/chapter/10.1007/3-540-63104-6_39
+
 \bibliography{paper}

### 2025-02-22 paper.tex
LLMs and direct manipulation

@@
 \newcommand{\vect}[1]{\lbrack #1 \rbrack}
-%\newcommand{\preview}{\textnormal{\ident{preview}}}
-%\newcommand{\links}{\textnormal{\ident{links}}}
+\newcommand{\preview}{\textnormal{\ident{preview}}}
+\newcommand{\links}{\textnormal{\ident{links}}}
 \DeclareMathOperator{\argmax}{arg\,max}
@@
 \begin{abstract}
-
-Some of the most remarkable results in mathematics reveal connections between different
+\noindent
+TODO: Some of the most remarkable results in mathematics reveal connections between different
 branches of the discipline. The aim of this paper is to point out a modest, but still
@@
 
-use a simple formal mathematical model
-that we call \emph{the choose-your-own-adventure calculus} to
-
+\noindent
 TODO: Say somewhere that this also formalizes things like mixed-initiative interaction and
@@
 
-% and enable a transfer of ideas between them
+~
+
+\noindent
+TODO: It is mainly about giving us a nice way to talk about lots of things in interactive
+programming systems and enable a transfer of ideas between them
 
@@
 analyst can easily spot the mistake. They click the ``+'' button to add a constraint and choose
-\ident{Don't match Urban.rural and Nation} to specify that the two columns should not be matched.
+\ident{Don't match `Urban.rural' and `Nation'} to specify that the two columns should not be matched.
 Datadiff then runs again and finds the correct matching.
@@
 \begin{itemize}
+\setlength{\itemsep}{5pt}
 \item A system for type-directed program synthesis could automatically synthesize parts of a
@@
 
-\subsection{LLM assistant for data exploration}
-
-\subsection{Programming by demonstration}
-
-ways of integrating automation with manual interaction
-- confirm each step vs. auto and retrace
-
-\cite{rein-2019-exploratory}
-
-\cite{verter-2024-mixed}
+\newpage
+
+\subsection{Language model-based completion}
+Large language models can assist with data exploration by generating snippets of code based
+on natural language description of the problem \cite{yin-2023-codegen}. A recognized drawback of
+this approach is that the user may gain only cursory understanding of the code and overlook
+errors. Various systems address this by generating explanations alongside with code \cite{nooralahzadeh-2024-data}.
+We developed a system that provides LLM-based natural language assistance for The Gamma [report
+citation omitted] based on a mechanism that assists the user and is also applicable to other
+choose-your-own-adventure systems.
+
+\begin{figure}
+{\raggedright\ttfamily\small
+You are helping user to complete a task in an interactive programming environment. The user's
+query is: "Give me the athlete with the largest number of gold medals."
+
+~\\[-0.75em]
+
+The query built so far is: "olympics"."group data"."by Athlete".
+
+~\\[-0.75em]
+
+The environment offers the user possible options. Choose an option that the should be applied to the current dataset:
+
+~\\[-0.75em]
+
+1.\hspace{0.4em} count distinct Athlete\\
+2.\hspace{0.4em} count distinct Discipline\\
+\hspace{2.3em}\lbrack\textit{multiple options omitted}\rbrack \\
+13. count all\\
+14. concatenate Athlete\\
+15. concatenate Discipline\\
+\hspace{2.3em}\lbrack\textit{multiple options omitted}\rbrack\\
+23. sum Gold\\
+24. sum Silver\\
+25. sum Year\\
+
+~\\[-0.75em]
+
+You should answer with the number of the option and no further explanation.}
+\caption{A prompt to complete a data exploration query based on a natural language
+question, which uses LLM to choose from the available completion options.}
+\label{fig:llm}
+\end{figure}
+
+\subparagraph{LLM assistant for The Gamma.}
+Our system [reprot citation omitted] integrates a type provider like the one used in The Gamma
+with a large language model (GPT-3.5 and GPT-4). It lets user ask a natural language query and
+then recommends choices that the user should make in order to answer the query. Our system does
+not use LLM to generate code, but to recommend an option offered by the type provider. To do so,
+we construct a prompt as shown in Figure~\ref{fig:llm}, which asks the LLM to recommend the next
+choice. In this example, the LLM is able to reliably respond with ``23'', which is the correct
+choice. Our system then pre-selects it in the completion list offered to the user.
+
+In contrast to one-shot generation of Python or SQL code snippets, our approach guides the user
+through the program construction, providing them with an opportunity to review and check if the
+program logic matches their expectations.
+
+In addition to the basic prompting strategy outlined above, we evaluated a strategy where the LLM
+is provided with lookahead information, i.e., for each of the choices, we also include a list of
+the choices that will available subsequently (formatted as either inline information in parentheses
+or as a nested bullet-point list). We use two different data sources to evaluate the system,
+each with 10 queries for which we manually determine the correct choices. To compare the results,
+we compute a score for each query by following the correct path, asking for an LLM recommendation
+in each step and computing the ratio of correct choices. Figure~\ref{fig:methods} shows the
+average scores for five different strategies, using two different LLM engines.
+
+\begin{figure}[t]
+\includegraphics[width=0.7\textwidth]{fig/llm.png}
+\caption{A plot showing the average scores for two different large language models (GPT-3.5 and GPT-4)
+solving 20 challenge problems using five different prompting strategies.}
+\label{fig:methods}
+\end{figure}
+
+The results suggest that, when using a more advanced LLM system, the ratio of correct
+recommendations is high-enough to be practically useful. It also shows that providing more
+information to the LLM through the lookahead mechanism improves the quality of the recommendation,
+in a consistent way across multiple LLM systems. Arguably, offering a recommendation is preferrable
+over providing an opaque solution, because it keeps the human in the loop \cite{sellen-2024-copilot}
+and avoids ``ironies of automation'' \cite{bainbridge-1983-ironies} including deskilling.
+
+\subparagraph{Generalised LLM assistant.}
+The LLM-based recommendation engine developed for The Gamma can be implemented for any
+choose-your-own-adventure system. As can be seen in Figure~\ref{fig:llm}, the information needed
+to construct an LLM prompt comprise only the natural language query entered by the user,
+a sequence of previously made choices $\iota_1, \ldots, \iota_{n-1}$ and the identifiers
+$\iota_n, \iota_n', \iota_n'', \ldots$ of the choices offered by the $\choices$ operation.
+As the LLM-based recommendations are based on natural language analysis of the prompt,
+the quality of the recommendations depends on how semantically meaningful the generated
+identifiers are.
+
+An LLM-based recommendation engine can be useful for a number of the choose-your-own-adventure
+interactive systems discussed in this paper:
+
+\begin{itemize}
+\setlength{\itemsep}{5pt}
+\item We demonstrated the usefulness of the system for data exploration in The Gamma.
+  Type providers for structured data \cite{petricek-2016-fsdata} typically generate small number
+  of choices that the user can navigate without assistance, but navigating the schemas generated by
+  type providers for semantic knowledge bases \cite{syme-2012-inforich} may be simplified through a
+  natural language query.
+\item The datadiff AI assistant discussed in Section~\ref{sec:motivation} matches columns based
+  on statistical distribution and ignores column names. An LLM-based recommendation engine is
+  useful in this case, because it is able to recommend the option \ident{Don't match
+  `Urban.rural' and `Nation'} based solely on its name. For other AI assistants, the LLM-based
+  recommendation engine would need access to the input data in order to be useful.
+\item The problem of predicting steps in order to construct a proof is known as
+  proofstep generation in literature focused on deep learning for theorem
+  proving~\cite{zhaoyu-2024-deep}. Although most systems use models trained specifically for the
+  task, there is also interest in using general-purpose LLMs~\cite{zhang-2023-proofs}.
+  Our approach suggests a potential prompting strategy.
+\end{itemize}
+
+It is interesting to note that the choose-your-own-adventure calculus provides a common structure
+that makes it possible to integrate multiple different approaches to assisting human in program
+construction, ranging from automatic search (if we can provide a suitable heuristic) to
+the use of LLMs (if generated choices carry enough semantic information). It can also serve as the
+basis for visual programming environments discussed next.
+
+\begin{figure}[t]
+  \includegraphics[width=0.9\textwidth]{fig/histogram.png}
+  \caption{Programming by demonstration in Histogram -- (1) the user constructs program to load
+    data, (2) then they filter data using graphical interface, which (3) records an operation
+    corresponding to the interaction in code.}
+  \label{fig:histogram}
+\end{figure}
+
+\subsection{Direct manipulation}
+
+The Histogram programming environment~\cite{petricek-2019-histogram} provides a code completion
+mechanism that can be used both to choose operations (as with type providers) and to specify
+their parameters (by offering available values in context as possible arguments). It makes it
+possible to evaluate sub-expressions at a fine-grained level and also refines type information
+based on runtime values. However, it also supports programming by demonstration
+\cite{cypher-1993-pbd,kandel-2011-wrangler}.
+
+As illustrated in Figure~\ref{fig:histogram}, when the user loads data, they can manipulate it
+in a table through a direct manipulation interface~\cite{shneiderman-1983-direct}. In Histogram,
+interacting with the table triggers a sequence of operations that construct code. A similar
+functionality is available in The Gamma, where operations for manipulating tabular data can be
+selected not only through auto-completion, but also by interacting with the live preview.
+
+The choose-your-own-adventure calculus enables a more general perspective on the programming
+by demonstration interaction as implemented by Histogram. When interacting with a live
+preview, the user is using a graphical interface to choose an option offered by $\choices$.
+The operation appends an item to the sequence of choices, updates the system state
+and generates a new preview where the choice is reflected (e.g., by filtering the table).
+
+An interesting question is whether a program can be constructed solely through direct manipulation,
+by interacting with the live previews. For this to be possible, the live preview needs to
+offer links to all the possible options offered by $\choices$. This requires a careful design
+of live previews. In Histogram (Figure~\ref{fig:histogram}), the live preview offers interactive
+elements for filtering based on equality test (row of drop-downs), sorting (up/down arrows)
+and indexing (indices in the ``\#'' column), but some operations (such as aggregation) can only
+be constructed through code. Previews in The Gamma provide interactive elements for all options,
+but those cannot always be integrated with the data display. For example, after choosing a grouping
+key (Figure~\ref{fig:groupby}), the preview shows the key column, but other columns are hidden.
+Specifying an aggregation thus requires an additional (somewhat arbitrary) ``+'' button that
+provides access to the available choices.
+
+\begin{figure}[t]
+\includegraphics[width=0.6\textwidth]{fig/thegamma3.png}
+\caption{Specifying aggregation in The Gamma through a live preview. Grouping keys can be
+  selected from a drop-down, while aggregations have to be added through a menu.}
+\label{fig:groupby}
+\end{figure}
+
+We can use the choose-your-own-adventure calculus to make the question whether all
+programs can be constructed through direct manipulation more precise. Assume
+$\preview(e)=p$ is a preview constructed for an expression $e$ and $\links(p)=\iota_1, \ldots, \iota_n$
+are identifiers that can be invoked through the preview (akin to links in a hypertext document).
+
+\begin{definition}[Direct manipulation]
+A choose-your-own-adventure system with previews defined by $\preview$ and $\links$ supports
+full direct manipulation if:
+\begin{itemize}
+\item $\forall \sigma, e$ such that $\select(\sigma)=e$ it is the case that
+  $\forall \iota'\mapsto\sigma'\in\choices(\sigma)\,.\,\iota'\!\in\!\links(\preview(e))$.
+\end{itemize}
+\end{definition}
+
+A system supports full direct manipulation if any program can be constructed by interacting with
+the live previews, created based on the gradually constructed programs. As illustrated by The Gamma,
+this can always be achieved by listing the options in a menu, but it is more interesting to see
+if the links can be directly embedded in the preview as in the data table interface of Histogram.
+
+The idea of directly mapping elements in a user interface to underlying structure of the
+programming system has previously been implemented in pioneering user interface systems
+including the Alternate Reality Kit~\cite{randall-1986-ark} and the Morphic framework for
+Self and Squeak \cite{maloney-1995-morphic,maloney-2001-morphic}. In those systems, user interface
+interactions directly map to messages sent to the underlying object and the halo element in
+Morphic plays a similar role to the additional menu in The Gamma. It remains to be seen if
+the choose-your-own-adventure calculus can provide a new way of looking at those systems.
 
@@
 
-hard to edit - you cannot easily go back
+TODO: Is this too general to be practically useful? E.g. LLM for interactive theorem provers
+arguably needs more specific information.
+
+~
+
+\noindent
+TODO: What if the list of choices is infinite?
+
+~
+
+\noindent
+TODO: How do you go back and revisit earlier choices? Backtracking in theorem provers or
+fixing earlier part of the query in data exploration. Sometimes, you can retrace the
+steps, but they may change...
+
+~
+
+\noindent
+Related work - mbark for emacs, VS code command palette
+
+~
+
+\noindent
+Alphageometry paper - follows a similar way of working to choose-your-own-adventure
+
+\section{Conclusions}
+
+TODO: Some conclusions
+1) lets us ask all those questions
+2) let ideas cross between domains
 
@@
 
-% cite: https://link.springer.com/chapter/10.1007/3-540-63104-6_39
-
 \bibliography{paper}

### 2025-03-05 paper.tex
Line numbers
Structure editors, some renaming

@@
 \hideLIPIcs
-
 \newcommand{\ident}[1]{\textsf{#1}}
@@
 \keywords{Interactive programming systems, Type providers, Proof assistants}
-\nolinenumbers
+%\nolinenumbers
 
@@
 
+\begin{figure}[t]
+  \includegraphics[width=0.55\textwidth]{fig/sandblocks.png}
+  \caption{Constructing a program using the Sandblocks structure editor \cite{beckmann-2023-all}.
+    The user is typing the start of an expression and context menu offers possible production
+    rules according to a grammar.}
+  \label{fig:sand}
+\end{figure}
+
+\subparagraph{Structure editors.}
+Structure editors make it possible to construct programs by manipulating the abstract syntax tree
+of a program rather than by working with textual representation. The interaction with a structure
+editor can be based on menus \cite{teitelbaum-1981-synth}, key bindings \cite{weber-2017-editors},
+tiles~\cite{moon-2022-tylr} and a range of other approaches.
+
+Sandblocks~\cite{beckmann-2023-all} is an example of a recent structure editor that is interesting
+for multiple reasons. First, the editor is automatically generated from a grammar. Second, the
+editor combines keyboard input with context menus (Figure~\ref{fig:sand}). In the above example,
+the user is typing code and needs to fill a placeholder for an expression. After they type the
+start of the expression, the editor offers a completion list with possible production rules from
+the grammar that are valid in the given context and are compatible with the text typed so far. The
+user can continue typing (to further refine the recommendation), but they can also select a rule
+from the menu. This completes the expression according to the production rule, leaving further
+empty holes in the code.
+
+In Sandblocks, the context menus serve more as hints and users of the editor often
+construct programs through typing, but the system illustrates the fact that structure editors
+can also be based on the interaction pattern captured in this paper. The editor can offer
+all possible production rules based on the given grammar and the user could construct an
+entire program by choosing one of the rules (and manually entering primitive terminals such as
+strings and identifiers).
+
+\newpage
+
 \section{Formal model}
@@
 output data) but is structurally similar. It is also worth noting that the definition may describe
-not just trees, but also graphs with cycles -- a system can return to an already visited state.
+not just trees, but also graphs with cycles -- a system can return to a previously visited state.
 This is not practically useful, but it does not pose a theoretical problem.
 
-\subparagraph{External mode of embedding.}
+\subparagraph{Expression completion.}
 One of the subtle questions about the choose-your-own-adventure pattern raised in the introduction
-concerns the different ways in which a trace of the interaction is embedded in the interactively
-constructed program. In the \emph{external mode}, the interaction results in code that becomes a
-part of the edited program, but it is not possible to reconstruct the steps used to generate the code.
+is whether a trace of the interaction can be reconstructed from the source code of an interactively
+constructed program. That is, if we follow a sequence of choices $\iota_1, \ldots, \iota_n$ to
+construct a program $e$, is it possible to recover the original sequene of choices from the
+program $e$.
 
@@
 To model this, we assume that the host language has a notion of a hole, written as~$?$
-and that a user can select a part of program to invoke the completion on. We write $E[e]$ for a
-completion context, akin to evaluation contexts in operational semantics.
-
+and that a user can select a part of program to invoke the completion on. When invoked on
+an expression containing multiple holes, the system can start by completing the first hole.
+
+We write $E[e]$ for a completion context, akin to evaluation contexts in operational semantics.
 We assume that, for a program containing a hole in a completion context $E[?]$, we can construct
-an initial choose-your-own-adventure state using an operation
-$\ident{init}(E[?])=\sigmaN$.
-
-\begin{definition}\label{def:external}
-An expression $E[?]$ is completed as $E[e]$ via external embedding of an interaction with
+an initial choose-your-own-adventure state using an operation $\ident{init}(E[?])=\sigmaN$.
+
+\begin{definition}[Expression completion]
+An expression $E[?]$ is completed as $E[e]$ through an interaction with
 a choose-your-own-adventure system consisting of \choices\ and \select\ if:
@@
 
-As we will see when we revisit the earlier examples formally, the external mode of embedding is
-used, for example, in the case of interactive theorem provers like \textsc{Alf} or Idris.
+As we will see when we revisit the earlier examples formally, this way of invoking a choose-your-own-adventure
+system is used, for example, in the case of interactive theorem provers like \textsc{Alf} or Idris.
 In those systems, the user triggers the completion on a proof (program) containing a hole.
@@
 
-\subparagraph{Internal mode of embedding.}
-In the \emph{internal mode}, a trace of the interaction with a choose-your-own-adventure system
-is embedded directly in the constructed program. This is the case with type providers, where a
-user chooses a sequence of object members to be accessed. The same would be the case in a completion
-system for Coq that would offer tactics to apply, becuase the resulting proof would contain a
-record of the selected tactics.
-
-To talk about the internal mode formally, we again need the \ident{init} operation, but also an
-operation \ident{decode} that extracts identifiers of invoked completions from an expression.
-An internal embedding is the same as external embedding with an additional constraint:
-
-\begin{definition}\label{def:internal}
-An expression $E[?]$ is completed as $E[e]$ via internal embedding of an interaction with
+\subparagraph{Reconstructible completion.}
+In many choose-your-own-adventure systems, it is possible to reconstruct a trace of the interaction
+through which a program was constructed. This is the case with type providers, where a user chooses
+a sequence of object members to be accessed and those members directly appear in the source code.
+The same would be the case in a completion system for Coq that would offer tactics to apply,
+becuase the resulting proof would contain a record of the selected tactics.
+
+For systems such as type providers, we say that that expression completion is \emph{reconstructible}.
+To capture the notion formally, we again need the \ident{init} operation, but also an operation
+\ident{decode} that extracts identifiers of invoked completions from an expression:
+
+\begin{definition}[Reconstructible expression completion]
+An expression $E[?]$ is reconstructibly completed as $E[e]$ through an interaction with
 a choose-your-own-adventure system consisting of \choices\ and \select\ if:
@@
 \begin{enumerate}
-\item $E[?]$ is completed as $E[e]$ via external embedding using \ident{init}, \choices\ and \select\\
+\item $E[?]$ is completed as $E[e]$ using \ident{init}, \choices\ and \select\\
   through a series of choices designated by identifiers $\iota_1, \ldots, \iota_n$,
@@
 
-If a choose-your-own-adventure system is integrated in a programming language
-through internal embedding, we can reconstruct the choices through which the user constructed
+If the integration of a choose-your-own-adventure system with a host programming language
+uses reconstructible expression completion, we can recover the choices through which the user completed
 an expression $e$ in a completion context $E[e]$, assuming they used the interactive system rather
@@
 \newpage
+\begin{figure}[t]
+\small
+(a) Grammar of a minimal language with declarations and expressions
+\[
+\begin{array}{rcl}
+\mathit{expr} & := & \mathit{literal} ~~|~~ \mathit{expr}+\mathit{expr} ~~|~~ \mathit{ident} ~~|~~ \mathit{expr}(\mathit{expr}^{*})\\
+\mathit{decl} & := & \kvd{function}~\mathit{ident}(\mathit{ident}^{*}) \{ \mathit{block} \} ~~|~~ \kvd{let}~\mathit{ident}=\mathit{expr}\\\
+\mathit{stmt} & := & \mathit{expr} ~~ | ~~ \mathit{decl}\\
+\mathit{block} & := & \mathit{stmt}^{*}
+\end{array}
+\]
+\begin{minipage}[t]{.5\textwidth}
+(b) Construction based on the grammar
+
+\begin{enumerate}
+\setlength{\itemsep}{0.75em}
+\item Add $\mathit{decl}$ as the first $\mathit{stmt}$ and fill two\\
+  $\mathit{ident}$ leafs (name and parameters)\\[0.5em]
+  $\begin{array}{l}
+  \kvd{function}~\ident{sayHello}(\ident{who})~\{\\
+  \quad\mathit{stmt}^{*}\\
+  \}\\
+  \mathit{stmt}^{*}
+  \end{array}$
+
+\item Sequentiall fill holes with invocations\\[0.5em]
+  $\begin{array}{l}
+  \kvd{function}~\ident{sayHello}(\ident{who})~\{\\
+  \quad\ident{print}(\texttt{''Hello ''} + \ident{who})\\
+  \}\\
+  \ident{sayHello}(\texttt{''world''})
+  \end{array}$
+\end{enumerate}
+\end{minipage}%
+\begin{minipage}[t]{.5\textwidth}
+(c) Alternative sequence based on refactorings
+
+\begin{enumerate}
+\setlength{\itemsep}{0.75em}
+\item Start with a small working program\\[0.5em]
+  $\begin{array}{l}
+  \ident{print}(\texttt{''Hello world''})
+  \end{array}$
+
+\item Extract literal into a variable\\[0.75em]
+  $\begin{array}{l}
+  \kvd{let}~\ident{who} = \texttt{''world''}\\
+  \ident{print}(\texttt{''Hello ''} + \ident{who})
+  \end{array}$
+
+\item Extract function with a parameter\\[0.75em]
+  $\begin{array}{l}
+  \kvd{function}~\ident{sayHello}(\ident{who})~\{\\
+  \quad\ident{print}(\texttt{''Hello ''} + \ident{who})\\
+  \}\\
+  \ident{sayHello}(\texttt{''world''})
+  \end{array}$
+
+\end{enumerate}
+\end{minipage}
+\caption{Comparison of a naive grammar-based program construction (b) and more user-friendly
+  program construction through a sequence of refactorings (c).}
+\label{fig:grammar}
+\end{figure}
+
+\subsection{Structure editors}
+Structure editors form a broad range of systems that support different kinds of interactions
+for constructing programs. This allows us to demonstrate some of the design considerations for
+choose-your-own-adventure systems. We first look at a simple choose-your-own-adventure system
+derived from a context-free grammar, which is inspired by Sandblocks~\cite{beckmann-2023-all}.
+We then discuss how such system could support refactorings in a way similar to
+\textsc{Deuce}~\cite{hempel-2018-deuce}, extending the basic grammar-directed way of working.
+
+\subparagraph{Formal model.}
+We model a system that, similarly to Sandblocks, generates choices automatically based on a given
+context-free grammar. We assume the grammar $(\mathcal{N},\Sigma,\mathcal{P},S)$ consists of set of
+non-terminal symbols $\mathcal{N}$, terminal symbols $\Sigma$, start symbol $S\in\mathcal{N}$
+and production rules $\mathcal{P}$ of the form $n\mapsto\boldsymbol{s}$ where
+$n\in\mathcal{N}$ and $\boldsymbol{s}\in(\mathcal{N}\cup\Sigma)^{*}$ (we write $n$ for non-terminals,
+$t$ for terminals and $s$ for all symbols; bold indicates a sequence of symbols).
+
+When interacting with the choose-your-own-adventure system, the user is gradually constructing a
+program (sentence) of a form $\boldsymbol{s}\in(\mathcal{N}\cup\Sigma)^{*}$. The system state is the program (sentence)
+constructed so far, i.e., $\sigma = \boldsymbol{s}$. The process can continue
+as long as there are non-terminal symbols in the sentence and production rules applicable to them.
+
+In general, the $\choices$ operation could recommend applicable production rules for any
+non-terminal in the sentence. The following always chooses the first non-terminal by assuming
+a sentence $\boldsymbol{t}n\boldsymbol{s}$ consisting of a sequence of terminals, followed by a
+single non-terminal and a sequence of arbitrary symbols (we discuss sequentialization
+in Section~\ref{sec:limitations}). The $\choices$ operation then finds all applicable production
+rules and offers them as choices. In a real-world system, identifiers $\iota_i$ are formed by
+the names of the rules (omitted from our formalism).
+\[
+\begin{array}{rcl}
+\choices(\boldsymbol{t}n\boldsymbol{s}) &=& \{\; \iota_i \mapsto \boldsymbol{t}\boldsymbol{s_i}\boldsymbol{s} ~|~ \forall (n\mapsto\boldsymbol{s_i})\in \mathcal{P} ~\}\\
+\select(\boldsymbol{s}) &=& \boldsymbol{s}
+\end{array}
+\]
+Unlike in the case of data exploration, the $\select$ operation does not ensure
+that a program is fully constructed and the returned sentence may contain further non-terminals.
+
+\subparagraph{Program refactorings.}
+The formal model derived from a context-free grammar is an adequate model for structure
+editors such as Sandblocks~\cite{beckmann-2023-all}, but many structure editors offer more
+advanced operations for program construction. \textsc{Deuce}~\cite{hempel-2018-deuce},
+which is a structure editor for a simple programming environment for creating Scalable
+Vector Graphics (SVG) images, for example, makes it possible to transform program through
+a range of refactorings such as extract function, introduce a local variable, inline definition,
+reorder arguments and more.
+
+Figure~\ref{fig:grammar} illustrates program construction via refactoring using a simple
+example. It defines a small language with a grammar. Figure~\ref{fig:grammar} (b) shows some
+of the steps necessary to construct a simple program via the grammar rules.
+Figure~\ref{fig:grammar} (c) shows a more appealing alternative where the programmer starts
+with a simple program and then gradually makes it more general through a sequence of refactorings.
+
+Modelling refactorings formally requires a more expressive framework~\cite{steimann-2011-refactoring},
+so we omit the details. The $\choices$ operation would need to identify applicable refactorings
+for the current program and offer them to the user.
+
+An important point illustrated by the two examples is whether the expression completion
+implemented by the two sketched choose-your-own-adventure systems is reconstructible.
+For a system based solely on context-free grammar, this is the case if the grammar is not
+ambiguous. Although the constructed expression does not directly contain an encoding of the
+sequence of operations, it is possible to reconstruct which production rules have been applied
+(by parsing the expression according to the grammar). A choose-your-own-adventure structure
+editor that supports refactorings no longer has this property, because it always offers multiple
+ways of constructing a given program (a program can be constructed with or without a refactoring).
+
+\newpage
 
@@
   The system is only complete if it is possible to choose a constraint set $\boldsymbol{c}$ that
-  restrict the set of programs $E_\boldsymbol{c}$ to a single given program. This is the case
+  restrict the set of programs $E_{\boldsymbol{c}}$ to a single given program. This is the case
   only for some AI assistants (e.g., datadiff, which always offers constraints to map column to
@@
 seen as design trade-offs that designers should consider.
+
+\subparagraph{Uniqueness.}
 
@@
 \section{Limitations}
+\label{sec:limitations}
 
@@
 arguably needs more specific information.
+
+\noindent
+TODO: Something about unnecessary seqentialization
 

### 2025-03-06 paper.tex
Submitted version

@@
-\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
+\documentclass[anonymous, a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
 \hideLIPIcs
@@
 \noindent
-TODO: Some of the most remarkable results in mathematics reveal connections between different
+Some of the most remarkable results in mathematics reveal connections between different
 branches of the discipline. The aim of this paper is to point out a modest, but still
 remarkable, similarity between a range of different interactive programming systems.
-
-~
-
-\noindent
-TODO: Say somewhere that this also formalizes things like mixed-initiative interaction and
-programming by demonstration
-
-~
-
-\noindent
-TODO: It is mainly about giving us a nice way to talk about lots of things in interactive
-programming systems and enable a transfer of ideas between them
+%
+In many programming systems, the user can interactively construct a program by repeatedly
+triggering some kind of completion mechanism and choosing one of the offered options. This
+is the case with code editors for object-oriented languages (choosing object members), data
+exploration environments (choosing an applicable transformation), but also theorem provers
+(choosing an applicable tactic) and structure editors (choosing a grammar~rule).
+
+In this paper, we formally capture the essence of this interaction pattern through a small
+formal model called \emph{the choose-your-own-adventure calculus}. We show how a wide range
+of different systems fits the model. Looking at the examples through a common perspective
+reveals multiple subtle differences. To formally capture those, we characterise
+basic properties of choose-your-own-adventure systems, resembling those from other areas
+of programming language research, including correctness, completeness and uniqueness.
+%
+We further show how the choose-your-own-adventure calculus can be used as the basis for
+formally studying more advanced interaction patterns including mixed-initiative interaction,
+AI-based programming assistants and programming by demonstration.
+
+We strongly believe that interaction with programming systems deserves as much attention
+as the underlying programming languages. Our work is one step in this direction. It provides
+a way of talking about commonalities and subtle differences between multiple interactive
+programming systems and enables a transfer of ideas between interactive systems from very
+different domains.
 
@@
 Multiple interactive programming systems, ranging from code editors for object-oriented programming
-languages to data exploration systems and interactive proof assistants, exhibit a remarkably
-similar pattern of interaction. They offer the user, who can be a programmer, a data scientist
+languages to data exploration systems, interactive proof assistants and structure editors, exhibit a
+remarkably similar pattern of interaction. They offer the user, who can be a programmer, a data scientist
 or a proof writer, a range of choices that the user can select from in order to complete their
@@
 options that are valid at a given location, or it may list only a subset of the valid options.
-In some cases, it may also include incorrect options as, for example, in auto-completion
+In some cases, it may even include incorrect options as, for example, in auto-completion
 for dynamic languages \cite{frolich-2024-autocomplete}.
 
-The aim of this paper is paper is to formally capture the recurring interaction pattern:
+\begin{figure}[t]
+  \includegraphics[width=0.7\textwidth]{fig/worldbank.png}
+  \caption{F\# code editor showing completions offered by the World Bank type provider.}
+  \label{fig:worldbank}
+\end{figure}
+
+The aim of this paper is to formally capture the recurring interaction pattern:
 
@@
 \setlength{\itemsep}{5pt}
-\item We motivate the formalism by reviewing four different systems that implement a variation on the
-interaction pattern. These include type providers for data access in F\# \cite{syme-2013-inforich},
+\item We motivate the formalism by reviewing five different systems that implement a variation on the
+interaction pattern. These include type providers in F\# \cite{syme-2013-inforich},
 type providers for data exploration in The Gamma \cite{petricek-2022-thegamma,petricek-2017-dotdriven},
-AI assistants for semi-automated data wrangling \cite{petricek-2023-aias} and tooling for
-interactive proof assistants \cite{altenkirch-1994-alf,brady-2015-idris,verter-2024-mixed} (Section~\ref{sec:motivation}).
+AI assistants for semi-automated data wrangling \cite{petricek-2023-aias}, tooling for
+interactive proof assistants \cite{altenkirch-1994-alf,brady-2015-idris,verter-2024-mixed}
+and structure editing \cite{beckmann-2023-all} (Section~\ref{sec:motivation}).
 
@@
 \item The calculus allows us to make the aforementioned subtle differences precise. We define the
-notions of \emph{correctness} and \emph{completeness} for the choose-your-own-adventure calculus.
-To distinguish the different ways of embedding the interactions in the edited programs, we
-also formally define \emph{internal} and \emph{external} mode of system integration.
-
-\item We show that various programmer assistance tools, such as search and AI-based recommendations
-can be built on top of the primitives offered by the calculus, showing how the
-choose-your-own-adventure calculus supports of transfer of ideas across different kinds of
+notions of \emph{correctness}, \emph{completeness} and \emph{uniqueness} for the choose-your-own-adventure
+calculus (Section~\ref{sec:properties}). To understand the way in which the interaction is
+integrated in a programming environment, we formally define the notion of \emph{expression
+completion} and capture a specfic kind of \emph{reconstructible} completion (Section~\ref{sec:calculus}).
+
+\item We show that multiple programmer assistance tools, such as mixed-initiative interaction,
+AI-based assistants and programming by demonstration can be defined on top of the primitives
+offered by the calculus (Section~\ref{sec:applications}). This also illustrates how the
+choose-your-own-adventure calculus supports transfer of ideas across different kinds of
 interactive programming systems.
@@
 
-\newpage
-
-\begin{figure}[t]
-  \includegraphics[width=0.7\textwidth]{fig/worldbank.png}
-  \caption{F\# code editor showing completions offered by the World Bank type provider.}
-  \label{fig:worldbank}
-\end{figure}
+% ==================================================================================================
 
@@
 
-The following sections review four different instances of the choose-your-own-adventure
-interaction pattern. In all of those, an interactive editor offers the user some kind of~a~com\-pletion list during working with the system.
+We start by reviewing five instances of the interaction pattern. In all of them, an editor
+offers the user a completion list to choose from during working with the system.
+
+\begin{figure}[t]
+\vspace{-0.5em}
+\includegraphics[width=0.48\textwidth]{fig/thegamma1.png}~~
+\includegraphics[width=0.48\textwidth]{fig/thegamma2.png}
+\caption{Constructing a query in The Gamma. We count the number of gold and silver medals for
+each athlete and sort the data by the number of gold medals.}
+\vspace{-0.5em}
+\label{fig:thegamma}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
 
@@
 
-F\# type providers \cite{syme-2013-inforich} are a mechanism for integrating external data
+F\# type providers \cite{syme-2013-inforich} are a mechanism for integrating (primarily) external data
 sources into the F\# type system. A type provider is a compiler extension, loaded and
@@
 external data and use it to generate a suitable statically-typed representation of the
-data, typically as objects with members. Type providers can, for example, infer the type from a
+data, typically objects with members. Type providers can, for example, infer the type from a
 sample JSON \cite{petricek-2016-fsdata} or read a database schema.
@@
 The result is a time series with values for the given indicator and a country. More generally,
-the example can be seen as a special case of a type provider for slicing n-dimensional
-data cube \cite{petricek-2022-thegamma} -- we choose a fixed value for two of the three
-dimensions (country, indicator, time).
-
-When using the type provider, the user types the first line of code and triggers auto-completion
-by typing \ident{wb} followed by the dot. The rest of the code is constructed by choosing an
-option from a list and typing another dot.\footnote{This interaction pattern has been
-lightheartedly called \emph{dot-driven development} by Phil Trelford \cite{seemann-2021-head}.}
-
-\newpage
-
-\begin{figure}[t]
-\includegraphics[width=0.48\textwidth]{fig/thegamma1.png}~~
-\includegraphics[width=0.48\textwidth]{fig/thegamma2.png}
-\caption{Constructing a query in The Gamma. We count the number of gold and silver medals for
-each athlete and sort the data by the number of gold medals.}
-\label{fig:thegamma}
-\end{figure}
+the example can be seen as a special case of a type provider for slicing $n$-dimensional
+data cube \cite{petricek-2022-thegamma}. We choose a fixed value for two of the three
+dimensions (country, indicator, time) and obtain a series indexed by the remaining dimension.
+
+When using the type provider, the user types the first line of code to initialize the type provider
+and triggers auto-completion by typing \ident{wb} followed by the dot. The rest of the code is
+constructed by choosing an option from a list and typing another dot (a pattern light-heartedly
+called \emph{dot-driven development} by Phil Trelford \cite{seemann-2021-head}).
+
+% --------------------------------------------------------------------------------------------------
 
@@
 repeatedly choosing operations and their parameters \cite{petricek-2017-dotdriven}. It keeps track
-of the schema and uses it to generate all possible valid parameters. When sortin data, it generates
-an object with two members for each columns -- one for ascending and one for descending sort.
+of the schema and uses it to generate all possible valid parameters. For example, when sorting data, it generates
+an object with two members for each columns -- one for ascending and one for descending order.
 Similarly, the grouping operation first offers all columns as possible grouping keys and then
 lets the user choose from a range of pre-defined aggregations (sum, count, average, concatenate).
-The system also evaluates the query on the fly, providing a live preview during editing \cite{petricek-2020-live}.
-
-The interaction pattern is the same as before. After the usertriggers auto-completion, they
+The system also evaluates the query on the fly, providing a live preview \cite{petricek-2020-live}.
+
+The interaction pattern is the same as before. After the user triggers auto-completion, they
 repeatedly select an operation and its parameters to construct a query. One notable difference
@@
 further operations) and so the types are generated lazily.
+
+\begin{figure}[t]
+  \vspace{-0.5em}
+  \includegraphics[width=0.875\textwidth]{fig/aia.png}
+  \caption{Using the datadiff AI assistant to reconcile the structure of the two datasets.
+    The user is offered a list of constraints to prevent or force matching between specific columns.}
+  \label{fig:aia}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
 
@@
 
-Figure~\ref{fig:aia} shows an example. It uses the datadiff \cite{sutton-2018-datadiff} AI
-assistant, running in a Wrattler notebook \cite{petricek-2018-wrattler}, to merge broadband
+Figure~\ref{fig:aia} shows an example. It uses the datadiff AI assistant \cite{sutton-2018-datadiff},
+running in a Wrattler notebook \cite{petricek-2018-wrattler}, to merge broadband
 quality data published by Ofcom for two subsequent years. The format of the CSV files for the
@@
 incorrectly maps the \ident{Urban.rural} (2014) column to \ident{Nation} (2015). This happens
-because both columns are categorical and have three values with similar distribution. A data
+because both columns are categorical and have three values with similar statistical distribution. A data
 analyst can easily spot the mistake. They click the ``+'' button to add a constraint and choose
@@
 Datadiff then runs again and finds the correct matching.
-
-\begin{figure}[t]
-  \includegraphics[width=0.9\textwidth]{fig/aia.png}
-  \caption{Using the datadiff AI assistant to reconcile the structure of the two datasets.
-    The user is offered a list of constraints to prevent or force matching between specific columns.}
-  \label{fig:aia}
-\end{figure}
-
-~
 
@@
 First, in the case of type providers, we are gradually constructing a program by adding operations to
-a method chain. Now, the AI assistant synthesizes a data transformation (program) and we are
+a method chain. An AI assistant automatically synthesizes a data transformation (program) and we are
 gradually adding constraints to control the synthesis. Second, in the case of type providers,
@@
 
-\newpage
-
 \begin{figure}[t]
+  \vspace{-0.5em}
   \includegraphics[width=0.6\textwidth]{fig/alf.png}
@@
 
+% --------------------------------------------------------------------------------------------------
+
 \subparagraph{Interactive theorem provers.}
@@
 also be offered through a user interface. An example that illustrates this is the interactive editor
-for the \textsc{Alf} theorem prover \cite{magnusson-1994-alf}, which is based on the refinement of
+for the \textsc{Alf} theorem prover \cite{magnusson-1994-alf}, which is based on a gradual refinement of
 an incomplete proof object \cite{altenkirch-1994-alf}. This is illustrated in Figure~\ref{fig:alf}.
@@
 The interaction pattern is again the same. The user repeatedly triggers a completion and uses
-it to refine and complete their proof by filling holes. There are subtle differences too. Unlike
-with AI assistants, each completion directly refines the proof that the user is editing. Unlike
-with type providers, a completion may generate multiple new holes, rather than just adding to a
+it to refine and complete a proof by filling holes. There are subtle differences too. Unlike
+earlier, each completion directly refines the proof that the user is editing;
+a completion may also generate multiple new holes, rather than just appending to a
 chain of operations.
@@
 
-\newpage
-
 \begin{figure}[t]
-  \includegraphics[width=0.55\textwidth]{fig/sandblocks.png}
+  \vspace{-0.25em}
+  \includegraphics[width=0.525\textwidth]{fig/sandblocks.png}
   \caption{Constructing a program using the Sandblocks structure editor \cite{beckmann-2023-all}.
@@
 
+% --------------------------------------------------------------------------------------------------
+
 \subparagraph{Structure editors.}
 Structure editors make it possible to construct programs by manipulating the abstract syntax tree
-of a program rather than by working with textual representation. The interaction with a structure
+of a program rather than by working with text. The interaction with a structure
 editor can be based on menus \cite{teitelbaum-1981-synth}, key bindings \cite{weber-2017-editors},
-tiles~\cite{moon-2022-tylr} and a range of other approaches.
+tiles~\cite{moon-2022-tylr} and a range of other approaches. A number of those fit
+the choose-your-own-adventure pattern.
 
@@
 user can continue typing (to further refine the recommendation), but they can also select a rule
-from the menu. This completes the expression according to the production rule, leaving further
-empty holes in the code.
-
-In Sandblocks, the context menus serve more as hints and users of the editor often
+from the menu. This completes the expression according to the production rule, generating further
+empty holes that can, in turn, be interactively filled with code.
+
+In Sandblocks, context menus serve mainly as hints and users of the editor often
 construct programs through typing, but the system illustrates the fact that structure editors
@@
 all possible production rules based on the given grammar and the user could construct an
-entire program by choosing one of the rules (and manually entering primitive terminals such as
+entire program by choosing one of the rules (and manually entering terminals such as
 strings and identifiers).
 
-\newpage
+% ==================================================================================================
 
@@
 We can think of the interaction with the system as navigating through a tree structure, starting
-from a root and choosing one of the possible branches in each step.\footnote{Serving as another
-evidence for the surprising effectiveness of the concept of a tree \cite{nesetril-2005-strom}.}
+from a root and choosing one of the possible branches in each step. %\footnote{Serving as another
+%evidence for the surprising effectiveness of the concept of a tree \cite{nesetril-2005-strom}.}
 In the following definition, the key $\choices$ operation can thus be seen as returning
@@
 The definition is not a programming language calculus in the usual sense in that it does not
-define a concrete syntax with reduction rules. It is an abstract algebraic structure that captures
+define a concrete syntax with reduction rules. It is an abstract algebraic description of
 the structure of a system that supports the choose-your-own-adventure interaction pattern.
@@
 output data) but is structurally similar. It is also worth noting that the definition may describe
-not just trees, but also graphs with cycles -- a system can return to a previously visited state.
-This is not practically useful, but it does not pose a theoretical problem.
+not just trees, but also graphs with cycles. A system where the user can return to a previously
+visited state may not be practically useful, but it does not pose a theoretical problem.
+
+% --------------------------------------------------------------------------------------------------
 
@@
 constructed program. That is, if we follow a sequence of choices $\iota_1, \ldots, \iota_n$ to
-construct a program $e$, is it possible to recover the original sequene of choices from the
-program $e$.
+construct a program $e$, is it possible to recover the original sequene of choices through which
+it was constructed just from the program $e$.
 
@@
 \item $\ident{init}(E[?]) = \sigmaN$ obtains the initial state of a choose-your-own-interaction system,
-\item $\sigma_n$ is a system state such that $\forall i\!\in\! 1\ldots n\;.\;(\iota_i\mapsto\sigma_i)\in\choices{(\sigma_{n-1})}$,
+\item $\sigma_n$ is a system state such that $\,\forall i\!\in\! 1\ldots n\;.\;(\iota_i\mapsto\sigma_i)\in\choices{(\sigma_{n-1})}$,
   i.e.,~the user\\ makes a series of choices resulting in a final state of the system $\sigma_n$,
-\item $E[e]$ where $e=\select(\sigma_n)$, i.e.,~the final program is constructed by replacing the hole
+\item $E[e]$ where $e=\select(\sigma_n)$ is the final program constructed by replacing the hole
   in the completion context with the expression $e$ generated from $\sigma_n$.
@@
 As we will see when we revisit the earlier examples formally, this way of invoking a choose-your-own-adventure
-system is used, for example, in the case of interactive theorem provers like \textsc{Alf} or Idris.
+system is used, for example, in the case of interactive theorem provers.
 In those systems, the user triggers the completion on a proof (program) containing a hole.
@@
 
+% --------------------------------------------------------------------------------------------------
+
 \subparagraph{Reconstructible completion.}
 In many choose-your-own-adventure systems, it is possible to reconstruct a trace of the interaction
-through which a program was constructed. This is the case with type providers, where a user chooses
+through which a program was constructed. This is the case with type providers, where the user chooses
 a sequence of object members to be accessed and those members directly appear in the source code.
 The same would be the case in a completion system for Coq that would offer tactics to apply,
-becuase the resulting proof would contain a record of the selected tactics.
+becuase the resulting proof would consist of a sequence formed by the selected tactics.
 
@@
 If the integration of a choose-your-own-adventure system with a host programming language
-uses reconstructible expression completion, we can recover the choices through which the user completed
-an expression $e$ in a completion context $E[e]$, assuming they used the interactive system rather
-than entering the code directly. This also means that we can reconstruct the final state $\sigma_n$
-of the system by starting from $\ident{init}(E[?])$ and following the choices specified by
-$\iota_1, \ldots, \iota_n$.
-
-\newpage
+uses reconstructible expression completion, we can recover the choices through which the user
+constructed the expression $e$ in a completion context $E[e]$. This also means that we can
+reconstruct the final state $\sigma_n$ of the system by starting from $\ident{init}(E[?])$
+and following the choices specified by $\iota_1, \ldots, \iota_n$.
+
+% ==================================================================================================
+
 \section{Examples}
@@
 
-We now revisit the four examples from Section~\ref{sec:motivation} and show how they fit the
-above formal model. All four examples rely on some domain-specific logic. We describe what
-information the logic provides, but do not model it formally. This has been done elsewhere,
-in works describing the individual systems.
+We now revisit the five examples from Section~\ref{sec:motivation} and show how they fit the
+above formal model. All five examples rely on some domain-specific logic. We describe what
+information the logic provides, but refer to full description elsewhere for details.
 
@@
 we start with a model of data exploration system that is inspired by The Gamma, but differs in
-one notable way. We then discuss type providers more generally and show how to correctly model
-The Gamma. We then revisit the remaining two examples.
+one notable way. We then discuss type providers more generally and show how to model The Gamma
+more precisely. We then revisit the remaining three examples.
+
+% --------------------------------------------------------------------------------------------------
 
@@
 members. The operation is encapsulated in the code of the accessor of the member. In the
-simplified model in this section we ignore this fact. The model presented here directly
+simplified model developed in this section we ignore this fact. The model presented here directly
 generates code that calls the underlying operations. For example, assume that the user makes
@@
 In The Gamma, the individual identifiers become object members and they are included as
-a member chain in the generated code. In the following simplified model, the completion
+a member chain in the generated code. In the simplified model presented here, the completion
 instead fills the hole with an expression representing the operation:
@@
 of mapping, while the former is less cognitively demanding to read for a non-programmer.
-As discussed in Section~\ref{sec:properties}, the two implementations of the choose-your-own-adventure
+As we will see in Section~\ref{sec:properties}, the two implementations of the
 interaction pattern also differ in terms of their formal properties.
@@
 the operation $\op_n$ and the already known parameters $\boldsymbol{p_n}$.
-The $\choices$ operation then generates options that add the additional parameter. We generated the
+The $\choices$ operation then generates options that add the additional parameter. We generate the
 identifiers $\iota',\iota'',\ldots$ based on the state and the parameter value, such as
@@
 and we generate identifiers $\iota',\iota'',\ldots$ based on the operation name. The $\ident{choices}$
-operation then returns options that add the additional operation:
+operation then returns options that append the additional operation:
 \[
@@
 possible operations and their parameters based on the current schema of the data. As the above
-definition shows, this underlying engine can be easily exposed through the common
+definition shows, this underlying engine can be wrapped and exposed through the common
 choose-your-own-adventure interface.
+
+% --------------------------------------------------------------------------------------------------
 
@@
 
-The type provider mechanism in F\# operates at the level of the type system. It is not a merely an
+The type provider mechanism in F\# operates at the level of the type system. It is not merely an
 editor feature. A type provider for a data source, such as the World Development Indicators database,
@@
 
-To model the completion mechanism, we also need to model information about types. We loosely
+To model the completion mechanism, we also need a representation of the type information. We loosely
 follow the Foo calculus model \cite{petricek-2016-fsdata} and write $\mathbb{C}$ for a set
@@
 Each member in the Foo calculus consists of a of a name $\iota$, return type $C$ and implementation
-$e$. For our purposes, we only need the type information and so the operations that define the
+$e$. For our purposes, we only need the type information and the operations that define the
 choose-your-own-adventure are parameterized by the set of classes $\mathbb{C}$.
@@
 The $\choices_\mathbb{C}$ operation finds the class definition corresponding to the type of the
-last member in the current chain. It offers choices appending each of the available
+last member in the current chain. It generates choices appending each of the available
 members to the current chain. The $\select_\mathbb{C}$ operation returns the constructed
 member chain:
-
 \[
@@
 
-The model follows the internal mode of embedding the interaction in the program. It is easy to
+The model follows the reconstructible expression completion model. It is easy to
 define the \ident{decode} operation that takes the resulting generated expression and returns the
@@
 As noted earlier, The Gamma does not embed query expressions directly into the generated code.
-It uses the same model as type providers and generates choices as members of
+It uses the model presented in this section and generates choices as members of
 types behind the scenes. We return to the differences between the two models in
 Section~\ref{sec:properties}.
+
+% --------------------------------------------------------------------------------------------------
 
@@
 different, way of implementing the choose-your-own-adventure pattern. The assistant operates
-with respect to some input data $X$ that does not change during the interaction and so we
+with respect to some input data $X$ that do not change during the interaction and so we
 parameterize the choose-your-own-adventure calculus operations by the data.
 The input data $X$ can be actual input data or a representative sample and so the AI assistant
-can be use past data to infer a cleaning script that will be used on new inputs.
+can use past data to infer a cleaning script that will be used on new inputs.
 
@@
 
+\vspace{-0.25em}
 \begin{itemize}
@@
 \end{itemize}
+\vspace{-0.25em}
 
@@
 the constraints and computes a sum of distances between the statistical distributions of the
-columns from $X_1$ and a patched version of $X_2$.
+columns from $X_1$ and the patched version of $X_2$.
 
@@
 human-readable description of the constraints. Note that this operation is specific to the particular
-AI assistant. The $\choices_X$ operation then offers a list of constraint sets where the additional
-constraint is added to the previously collected set:
+AI assistant; $\choices_X$ then offers a list of constraint sets with one of the additional constraints:
 \[
@@
 \]
-The integration of an AI assistant, as described here, has to follow the external mode of embedding.
-The interaction with the assistant results in a cleaning script (expression), but there is no way
-of reconstructing the constrains used to guide the optimization. To support internal embedding, the
+The expression completion for an AI assistant, as described here, is not reconstructible.
+The inter\-action results in a cleaning script (expression), but there is no way of reconstructing
+the constrains used to guide the optimization. To make the completion reconstructible, the
 $\select$ operation would need to explicitly include the constraints in the resulting expression.
@@
 
-\newpage
+% --------------------------------------------------------------------------------------------------
 
@@
 for dependently typed languages implement similar kinds of interactions, there is no well-documented
-system that exactly fits the pattern. The closest example is perhaps the recently envisioned
+system that fits the pattern exactly. The closest example is perhaps the recently envisioned
 mixed-mode interaction theorem prover \cite{verter-2024-mixed}. Rather than reframing the
@@
 term, or program, of a type that represents the theorem. Interactive editors exist for both
-types of systems. For Coq, the \texttt{Company-Coq} \cite{pitclaudel-2016-companycoq,aspinall-2000-general}
+kinds of systems. For Coq, the \texttt{Company-Coq} \cite{pitclaudel-2016-companycoq,aspinall-2000-general}
 extension offers auto-completion, which recommends available tactics, hypotheses and local
@@
 Despite the different ways of working, an implementation of the choose-your-own-adventure
-pattern for both types of systems would be similar. Based on the sub-goal that the user is
-currently proving, the system would recommend a range of tactics that can be applied to the
-sub-goal. In the case of Coq, the selected tactic would be added to the sequence. In Idris, the
-selected tactic would be applied to transform the current term. The difference is in the mode of
-embedding. A system for Coq would provide internal embedding in that the selected option is added
+pattern for both kinds of systems would be similar. Based on the current sub-goal, the system
+would recommend a range of tactics that can be applied to the sub-goal. In the
+case of Coq, the selected tactic would be added to the sequence. In Idris, the selected tactic
+would be applied to transform the current term. The difference is in the expression completion.
+A system for Coq would provide reconstructible completion in that the selected option is added
 to the proof source code. (Much like selecting a completion when using type providers appends a
-member access.) A system for Idris-like language would use the tactic to transform the term,
-making it impossible to reconstruct the sequence of applied tactics as in the external mode of
-embedding.
+member access.) A system for Idris or Agda would use the tactic to transform the term,
+making it impossible to reconstruct the sequence of applied tactics.
 
 \begin{figure}
+\vspace{-1.5em}
 \[
 \begin{array}{lcl}
-\boxed{\Gamma \vdash \tau \Rightarrow e}\\[1em]
+%\boxed{\Gamma \vdash \tau \Rightarrow e}\\[1em]
 \dfrac{}{\Gamma,x:\tau \vdash \tau \Rightarrow x}~\ident{(syn-var)} &\quad&
@@
 \]
+\vspace{-0.25em}
 \caption{Illustrative set of simple type-directed program synthesis rules}
 \label{fig:synths}
+\vspace{-0.25em}
 \end{figure}
@@
 interactive choose-your-own-adventure system. If the interaction was invoked to fill a typed
-hole $?_\tau$ in a context $\Gamma$, the system could collect multiple $e$ such that
-$\Gamma \vdash \tau \Rightarrow e$ and offer a choice of such options. Note that the definition
+hole $?_\tau$ in a context $\Gamma$, the system could collect all expressions $e$ such that
+$\Gamma \vdash \tau \Rightarrow e$ and offer a choice of those options. Note that the definition
 in Figure~\ref{fig:synths} synthesizes sub-expressions recursively, but a choose-your-own-adventure
-system may always fill those with a typed hole using (\ident{syn-hole}).
+system may proceed one step at a time, filling sub-expressions with typed holes using (\ident{syn-hole}).
 
@@
 completion context containing a (typed) hole; we assume that the variables $\Gamma$ available in
-the completion context of the hole can be obtained using $\ident{vars}(E[\_])$.
+the completion context of the hole can be obtained using $\ident{vars}(E[?_\tau])$.
 
@@
 \begin{array}{rcl}
-\choices(E[?_\tau]) &=& \{\; \iota \mapsto e ~|~ \forall e\;.\; \ident{vars}(E[\_]) \vdash \tau \Rightarrow e_i \}\\
+\choices(E[?_\tau]) &=& \{\; \iota \mapsto E[e] ~|~ \forall e\;.\; \ident{vars}(E[?_\tau]) \vdash \tau \Rightarrow e_i \}\\
 \select(e) &=& e
@@
 The $\choices$ operation synthesizes possible terms of a type required by the hole. Since the
-state $\sigma$ is a term, the $\select$ operation simply returns it. The definition models the
-external embedding of the interaction, i.e., a system that behaves according to Idris. It
-constructs the term, but does not record the completion choices. That said, a system akin to Coq
-that constructs a sequence of tactics could be modelled too if the state was a sequence of
-tactics and $\choices$ appended the available tactics as options to the end of the current sequence.
-
-\newpage
+state $\sigma$ is a term, the $\select$ operation simply returns it. The expression completion
+mode of the system behaves similarly to Idris. It constructs the term, but does not record the
+completion choices. A system akin to Coq that constructs a sequence of tactics could be modelled
+too if the state was a sequence of tactics and $\choices$ appended the available tactics
+to the end of the current sequence.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Structure editors}
+\label{sec:examples-struct}
+
+Structure editors implement a broad range of patterns of interaction for program construction.
+They allows us to demonstrate some of the design considerations for
+choose-your-own-adventure systems. We contrast a simple choose-your-own-adventure system
+derived from a context-free grammar, inspired by Sandblocks~\cite{beckmann-2023-all},
+with a system that extends the basic grammar-directed program construction with supports
+refactorings, akin to \textsc{Deuce}~\cite{hempel-2018-deuce}.
+
+\subparagraph{Formal model.}
+We model a system that, similarly to Sandblocks, generates choices automatically based on a given
+context-free grammar. We assume the grammar $(\mathcal{N},\Sigma,\mathcal{P},S)$ consists of set of
+non-terminal symbols $\mathcal{N}$, terminal symbols $\Sigma$, start symbol $S\in\mathcal{N}$
+and production rules $\mathcal{P}$ of the form $n\mapsto\boldsymbol{s}$ where
+$n\in\mathcal{N}$ and $\boldsymbol{s}\in(\mathcal{N}\cup\Sigma)^{*}$ (we write $n$ for non-terminals,
+$t$ for terminals and $s$ for all symbols; bold indicates a sequence of symbols).
+
+When interacting with the choose-your-own-adventure system, the user is gradually constructing a
+program (sentence) of a form $\boldsymbol{s}\in(\mathcal{N}\cup\Sigma)^{*}$. The system state is the program (sentence)
+constructed so far, i.e., $\sigma = \boldsymbol{s}$. The process can continue
+as long as there are non-terminal symbols in the sentence and production rules applicable to them.
+
+In general, the $\choices$ operation could recommend applicable production rules for any
+non-terminal in the sentence. The following always chooses the first non-terminal by assuming
+a sentence $\boldsymbol{t}n\boldsymbol{s}$ consisting of zero or more terminals, followed by a
+single non-terminal and a sequence of arbitrary symbols (the need to sequentialize the construction
+is a limitation discussed in Section~\ref{sec:limitations}). The $\choices$ operation then finds
+all applicable production rules and offers them as choices. In a real-world system, identifiers
+$\iota_i$ are formed by the names of the rules (omitted from our formalism).
+\[
+\begin{array}{rcl}
+\choices(\boldsymbol{t}n\boldsymbol{s}) &=& \{\; \iota_i \mapsto \boldsymbol{t}\boldsymbol{s_i}\boldsymbol{s} ~|~ \forall (n\mapsto\boldsymbol{s_i})\in \mathcal{P} ~\}\\
+\select(\boldsymbol{s}) &=& \boldsymbol{s}
+\end{array}
+\]
+Unlike in the case of data exploration, the $\select$ operation does not ensure
+that a program is fully constructed and the returned sentence may contain further non-terminals.
+
 \begin{figure}[t]
@@
 \]
+
 \begin{minipage}[t]{.5\textwidth}
-(b) Construction based on the grammar
+(b) Construction based on the grammar\\[-0.5em]
 
@@
 \begin{minipage}[t]{.5\textwidth}
-(c) Alternative sequence based on refactorings
+(c) Alternative sequence based on refactorings\\[-0.5em]
 
@@
 
-\subsection{Structure editors}
-Structure editors form a broad range of systems that support different kinds of interactions
-for constructing programs. This allows us to demonstrate some of the design considerations for
-choose-your-own-adventure systems. We first look at a simple choose-your-own-adventure system
-derived from a context-free grammar, which is inspired by Sandblocks~\cite{beckmann-2023-all}.
-We then discuss how such system could support refactorings in a way similar to
-\textsc{Deuce}~\cite{hempel-2018-deuce}, extending the basic grammar-directed way of working.
-
-\subparagraph{Formal model.}
-We model a system that, similarly to Sandblocks, generates choices automatically based on a given
-context-free grammar. We assume the grammar $(\mathcal{N},\Sigma,\mathcal{P},S)$ consists of set of
-non-terminal symbols $\mathcal{N}$, terminal symbols $\Sigma$, start symbol $S\in\mathcal{N}$
-and production rules $\mathcal{P}$ of the form $n\mapsto\boldsymbol{s}$ where
-$n\in\mathcal{N}$ and $\boldsymbol{s}\in(\mathcal{N}\cup\Sigma)^{*}$ (we write $n$ for non-terminals,
-$t$ for terminals and $s$ for all symbols; bold indicates a sequence of symbols).
-
-When interacting with the choose-your-own-adventure system, the user is gradually constructing a
-program (sentence) of a form $\boldsymbol{s}\in(\mathcal{N}\cup\Sigma)^{*}$. The system state is the program (sentence)
-constructed so far, i.e., $\sigma = \boldsymbol{s}$. The process can continue
-as long as there are non-terminal symbols in the sentence and production rules applicable to them.
-
-In general, the $\choices$ operation could recommend applicable production rules for any
-non-terminal in the sentence. The following always chooses the first non-terminal by assuming
-a sentence $\boldsymbol{t}n\boldsymbol{s}$ consisting of a sequence of terminals, followed by a
-single non-terminal and a sequence of arbitrary symbols (we discuss sequentialization
-in Section~\ref{sec:limitations}). The $\choices$ operation then finds all applicable production
-rules and offers them as choices. In a real-world system, identifiers $\iota_i$ are formed by
-the names of the rules (omitted from our formalism).
-\[
-\begin{array}{rcl}
-\choices(\boldsymbol{t}n\boldsymbol{s}) &=& \{\; \iota_i \mapsto \boldsymbol{t}\boldsymbol{s_i}\boldsymbol{s} ~|~ \forall (n\mapsto\boldsymbol{s_i})\in \mathcal{P} ~\}\\
-\select(\boldsymbol{s}) &=& \boldsymbol{s}
-\end{array}
-\]
-Unlike in the case of data exploration, the $\select$ operation does not ensure
-that a program is fully constructed and the returned sentence may contain further non-terminals.
-
 \subparagraph{Program refactorings.}
@@
 editors such as Sandblocks~\cite{beckmann-2023-all}, but many structure editors offer more
-advanced operations for program construction. \textsc{Deuce}~\cite{hempel-2018-deuce},
-which is a structure editor for a simple programming environment for creating Scalable
-Vector Graphics (SVG) images, for example, makes it possible to transform program through
+advanced operations for program construction. For example \textsc{Deuce}~\cite{hempel-2018-deuce},
+which is a structure editor for a programming environment for creating Scalable
+Vector Graphics (SVG) images, makes it possible to transform program through
 a range of refactorings such as extract function, introduce a local variable, inline definition,
@@
 Figure~\ref{fig:grammar} (c) shows a more appealing alternative where the programmer starts
-with a simple program and then gradually makes it more general through a sequence of refactorings.
+with a simple program and then gradually refines it through a sequence of refactorings,
+making it more general.
 
@@
 implemented by the two sketched choose-your-own-adventure systems is reconstructible.
-For a system based solely on context-free grammar, this is the case if the grammar is not
-ambiguous. Although the constructed expression does not directly contain an encoding of the
+For a system based on non-ambiguous context-free grammar, this is the case.
+Although the constructed expression does not directly contain an encoding of the
 sequence of operations, it is possible to reconstruct which production rules have been applied
@@
 editor that supports refactorings no longer has this property, because it always offers multiple
-ways of constructing a given program (a program can be constructed with or without a refactoring).
+ways of constructing a given program (a program can be constructed with or without a refactoring,
+or by using multiple refactorings in different orders).
 
@@
 programming systems interact with the user. We saw this in Section~\ref{sec:calculus}, which
-defines internal and external mode of embedding to distinguish between systems where the
+defines reconstructible expression completion to distinguish between systems where the
 interaction leaves a reconstructible trace in the constructed program and systems where it does not.
 In this section, we make precise two properties that were introduced informally in the context of
-data exploration in The Gamma \cite{petricek-2022-thegamma}.
+data exploration in The Gamma \cite{petricek-2022-thegamma} and an additional property related
+to reconstruction of interaction traces.
 
@@
 of correct expressions $\mathcal{E}$ to be hole-free, i.e., only programs that can run (or
-represent complete proofs) are correct. For systems where the completion is string-based,
-we may treat all syntactically-correct programs as correct.
+represent complete proofs) are correct. For systems where the completion is based on a grammar,
+we may require that correct programs do not contain non-terminals.
 
@@
 $\mathcal{E}$. Trivially, all systems are correct with respect to $\mathcal{E}=E$.
-However, the systems discussed in Section~\ref{sec:examples} are, with one exception,
+However, the three of the five systems discussed in Section~\ref{sec:examples} are also
 correct with respect to a non-trivial set of correct expressions:
@@
   are sound~\cite{osera-2015-synthesis}, that is if $\Gamma \vdash \tau \Rightarrow e$ then also
-  $\Gamma \vdash e : \tau$. Note that a correct choose-your-own-adventure system can be defined
+  $\Gamma \vdash e : \tau$. Note that a valid choose-your-own-adventure system can be defined
   even using unsound synthesis rules -- it would be sufficient to filter the recommended
   expressions in $\choices$ to the ones that are well-typed.
+
+\item A structure editor based on context-free grammar (Section~\ref{sec:examples-struct})
+  is not correct if we define $\mathcal{E}$ as the set of sentences that do not contain
+  non-terminals. A correct structure editor could be implemented if it was based
+  solely on refactorings such as those in Figure~\ref{fig:grammar}~(c).
 \end{itemize}
 
-There may be useeful systems that violate the correctness property. An tool based on a large
+There may be other useful systems that violate the correctness property. A tool based on a large
 language model (LLM) may generate code with errors that the programmer can later
@@
 The data exploration system discussed in Section~\ref{sec:examples-data} ensures correctness
-by dropping the last non-fully-specified operation in the $\select$ operation. As a result,
-it is correct, but it does not support internal embedding. If the operation is dropped, we cannot
-reconstruct the full sequence of interactions from the generated source code. Generating code
-that includes the non-fully-specified operation allows external embedding, but makes the system
-incorrect. It would still satisfy a weaker definition of (eventual) correctness.
+by dropping the last non-fully-specified operation in the $\select$ operation. As a result, it is
+correct, but it does not support reconstructible expression completion. If the operation is dropped,
+we cannot reconstruct the sequence of interactions from the source code. Generating code
+that includes the non-fully-specified operation allows reconstructibility, but makes the system
+incorrect. It would still satisfy a weaker eventual correctness property:
 
@@
 Eventual correctness models systems where some sequences of choices result in invalid programs,
-but it is always possible to make further choices to reach a valid program. It is
-always possible to turn an eventually correct system into a correct one:
+but it is always possible to reach a valid program. This includes
+structure editors based on context-free grammars, as well as the data exploration system.
+Interestingly, it is always possible to turn an eventually correct system into a correct one:
 
@@
   reached. This makes any eventually correct choose-your-own-adventure system correct, but it
-  does not support internal embedding.
+  breaks reconstructibility of expression completion.
 
@@
   identifiers $\iota_1,\ldots,\iota_n$.
-  This makes the system correct and also preserves external embedding, but it potentially
+  This makes the system correct and also preserves reconstructibility, but it potentially
   generates too many choices that are difficult to navigate.
@@
 
-There is more to be said about correctness of interactive programming systems, but the
+There is more to be said about correctness of interactive programming systems. The
 conceptual framework provided by the choose-your-own-adventure calculus makes it possible to
@@
 \begin{itemize}
-\item $\forall e\!\in\!\mathcal{E}\,.\,\exists \sigma_1,..,\sigma_n$ and $\iota_i,..,\iota_n$ such that
+\item $\forall e\!\in\!\mathcal{E}\,.\,\exists \sigma_1,..,\sigma_n$ and $\iota_1,..,\iota_n$ such that
   $\iota_i\mapsto\sigma_i \in \choices(\sigma_{i-1})$ and $e=\select(\sigma_n)$.
@@
 A system is complete if, for any correct program, there is a sequence of choices that can be
-used to construct the given program. This is a more subtle property than correctness and
-it does not hold for all the examples dicusssed in Section~\ref{sec:examples}.
+used to construct the given program. This is a more subtle property than correctness. It also
+does not hold for all the examples dicusssed in Section~\ref{sec:examples}.
 
@@
   parameter (sorting by a key, but not based on a custom expression). A completion system for a
-  more general-purpose query language would be incomplete.
+  more general-purpose query language, such as SQL, would be incomplete.
 
 \item For type providers (Section~\ref{sec:examples-tps}), the completion mechanism is complete,
-  because it offers all available members of the type (and ``.'' has to be followed by a member
-  access). Consequently, the type provider in The Gamma is also complete. Even if the underlying
-  query language is more expressive, it is hidden from the user and the system offers all
-  available members.
+  because it offers all available members of the type. Consequently, the type provider in The
+  Gamma is also complete. Even if the underlying query language is more expressive, it is hidden
+  from the user and the system offers all available members.
 
@@
   based on the current inferred program. It does not let the user construct arbitrary constraints.
-  Moreover, because the $\mathit{arg\,max}$ operation used in $\select$ performs statistical
+  Moreover, because the $\argmax$ operation used in $\select$ performs statistical
   optimization, there is no guarantee that it can be used to generate a specific program.
@@
   restrict the set of programs $E_{\boldsymbol{c}}$ to a single given program. This is the case
-  only for some AI assistants (e.g., datadiff, which always offers constraints to map column to
-  any chosen other column).
+  for some AI assistants, including datadiff, which always offers constraints to map column to
+  any chosen other column.
 
@@
   may introduce), but the great number of generated options would be impractical. A realistic
-  system would only generate a subset of most useful proof/program steps and let the user write
+  system would only generate a subset of the most useful proof/program steps and let the user write
   other steps manually (interactive proof construction in Idris can be seen as operating in this way).
+
+\item Completeness is easy to show for structure editors based on context-free grammars
+  (Section~\ref{sec:examples-struct}). For each sentence that can be produced by the rules of
+  the grammar, there is a sequence of choices that applies the necessary rules. However, editors
+  based on refactorings may not have the property as some programs may not be reachable just
+  through refactoring transformations.
 \end{itemize}
@@
 
+\newpage
+
 \subparagraph{Uniqueness.}
+The grammar-based structure editor example raised an interesting question about reconstructing
+interaction steps. Although the generated program does not directly have a sequential structure,
+it was possible to reconstruct the interaction steps from the expression (by parsing it and
+finding corresponding production rules). This is no longer possible if the program can be
+transformed through richer set of refactorings. The difference between the two examples is that
+the former has the uniqueness property:
+
+\begin{definition}[Uniqueness]
+A~choose-your-own-adventure system has the uniqueness property if and only if,
+for all expressions $e\in E$ it is the case that:
+\begin{itemize}
+\item if there are $\sigma_1,..,\sigma_n$ and $\iota_i,..,\iota_n$ such that
+  $\iota_i\mapsto\sigma_i \in \choices(\sigma_{i-1})$ and $e=\select(\sigma_n)$,
+\item and there are $\sigma'_1,..,\sigma'_n$ and $\iota'_i,..,\iota'_n$ such that
+  $\iota'_i\mapsto\sigma'_i \in \choices(\sigma'_{i-1})$ and $e=\select(\sigma'_n)$,
+\item then $\forall i\!\in\!\ 1\ldots n\,.\,\iota_i=\iota'_i\,\wedge\,\sigma_i=\sigma'_i$.
+\end{itemize}
+\end{definition}
+
+The uniqueness property guarantees that, for any program, there is only one way of constructing
+it. This is the case for the data exploration environment (Section~\ref{sec:example-data}) and
+type providers (Section~\ref{sec:examples-tps}) where the program directly mirrors the interactive
+steps. It is not the case for AI assistants (Section~\ref{sec:examples-aias}), where adding
+an irrelevant constraint may have no effect on the resulting program.
+
+A type-directed synthesis system or an interactive theorem prover (Section~\ref{sec:examples-thm})
+would have the property if no two tactics produce the same term. This is the case for our small
+example, but it may not hold in general. Finally, a structure editor based on an unambiguous
+context-free grammar (Section~\ref{sec:examples-struct}) satisfies uniqueness, but a structure
+editor that allows further refactorings does not, in general, does not.
+
+In theory, uniqueness guarantees that we can define the \ident{decode} operation required
+by reconstructible expression completion. However, uniqueness does not guarantee that there is
+a computationally effective way. To decode an expression efficiently, we need to be able to
+start from the initial state and identify the correct choice in each step. This is possible for
+a structure editor based on a context-free grammar. As a counter-example, consider a system that
+offers a large tree of choices and produces expression that is a hash of the choices made. For
+such system, there is only one path to each hash, but reconstructing it requires searching the
+entire tree. To guarantee a computationally effective reconstructibility, we thus need more than
+uniqueness -- in particular, it needs to be possible to uniquely determine what part of the
+constructed program corresponds to which of the choices from the sequence.
 
@@
 systems as instances of the same general pattern. This makes it possible to discuss properties
-of the systems (as done in the previous section), reuse components in system implementations,
+of the systems, reuse components in system implementations,
 but also transfer ideas across different domains. In this section, we discuss three ideas that
-emerged in the context of a specific interactive system, but could be applied to any system
-based on the choose-your-own-adventure pattern.
-
-~
+emerged in the context of a specific interactive system, but could be applied to other systems
+based on the pattern.
 
 \begin{figure}[t]
+\small
 \definecolor{shade1}{HTML}{f0f0f0}
@@
 \subsection{Mixed-initiative interaction}
-When using an interactive theorem prover, one typically constructs the proof manually until an
+When using a conventional interactive theorem prover, one typically constructs the proof manually until an
 automated strategy can fill in the remaining gaps. This is the case with Idris proof search,
@@
 As an illustration of the mixed-initiative proving, consider Figure~\ref{fig:proof} which
-shows a proof of commutativity of $+$ in Peano arithmetic. The figure shows a version of our example
-[report citation omitted],\footnote{Adapted from \url{https://github.com/boyland/sasylf/blob/master/examples/sum.slf}}
-simplified for brevity and written using the SASyLF notation \cite{aldrich-2008-sasylf}.
+shows a proof of commutativity of $+$ in Peano arithmetic. The figure shows a version of our example,
+adapted from SASyLF \cite{aldrich-2008-sasylf,aldrich-2025-sum} and simplified for brevity.
 After writing proofs of \ident{sum-z-rh} and \ident{sum-s-rh} (not shown), the user states
@@
 
-\subparagraph{Generalised mixed-initiative interaction.}
-The mixed-initiative mode of interaction combines manual interaction with automatic search.
-In order to support automatic search, the system needs a metric that determines whether a
-constructed program is correct (as when proving a given theorem) or whether it is an improvement
-over another equivalent program (for example when refactoring). This general way of working
-can be used in other interactive programming systems:
-
-\begin{itemize}
-\setlength{\itemsep}{5pt}
-\item A system for type-directed program synthesis could automatically synthesize parts of a
-  program (e.g., pattern matching and implementation for some of the cases), but ask user for
-  input in order to complete branches where solution was not found automatically.
-\item In data exploration, a system could perform automatic search through the available operations
-  in order to transform data into a more regular format, according to a suitability
-  metric as done, for example, in the Proactive Wrangling system~\cite{guo-2011-proactive}.
-\end{itemize}
-
-We can understand how mixed-initiative interactive systems as extensions
-built on top of the choose-your-own-adventure calculus. A mixed-initiative interactive system
-defines an operation \ident{suggest} that recommends a sequence of choices for a given state.
-
-\begin{definition}[Mixed-initiative system]
-A choose-your-own-adventure system supports mixed-initiative mode of interaction if it is
-equipped with an operation $\suggest$ such that:
-\begin{itemize}
-  \item $\suggest(\sigma_0)=\iota_1, \ldots, \iota_n$ such that
-    $\forall i\!\in\! 1\ldots n\;.\;\iota_i\mapsto\sigma_i\in\choices(\sigma_{i-1})$.
-\end{itemize}
-\end{definition}
-
-The definition only requires that the suggested sequence of choices is valid, but it
-does not specify how exactly the system should make the recommendations. This is specific to
-a particular system. An interactive theorem prover will try to find a proof or solve
-sub-goals, whereas data wrangling system may try to improve the structure of data.
-In practice, the suggestion should improve the quality of the constructed program so that
-$\select(\sigma_n)$ is better than $\select(\sigma_0)$, but we leave the definition flexible
-to accommodate a broader range of potential uses.
-
-\newpage
-
-\subsection{Language model-based completion}
-Large language models can assist with data exploration by generating snippets of code based
-on natural language description of the problem \cite{yin-2023-codegen}. A recognized drawback of
-this approach is that the user may gain only cursory understanding of the code and overlook
-errors. Various systems address this by generating explanations alongside with code \cite{nooralahzadeh-2024-data}.
-We developed a system that provides LLM-based natural language assistance for The Gamma [report
-citation omitted] based on a mechanism that assists the user and is also applicable to other
-choose-your-own-adventure systems.
-
-\begin{figure}
+\begin{figure}[t]
 {\raggedright\ttfamily\small
@@
 
+\subparagraph{Generalised mixed-initiative interaction.}
+The mixed-initiative mode of interaction combines manual interaction with automatic search.
+In order to support automatic search, the system needs a metric that determines whether a
+constructed program is correct (as when proving a given theorem) or whether it is an improvement
+over another equivalent program (for example when refactoring). This general way of working
+can be used in other interactive programming systems:
+
+\begin{itemize}
+\setlength{\itemsep}{5pt}
+\item A system for type-directed program synthesis could automatically synthesize parts of a
+  program (e.g., pattern matching and implementation for some of the cases), but ask user for
+  input in order to complete branches where solution was not found automatically.
+\item In data exploration, a system could perform automatic search through the available operations
+  in order to transform data into a more regular format, according to a suitability
+  metric as done, for example, in the Proactive Wrangling system~\cite{guo-2011-proactive}.
+\end{itemize}
+
+We can understand how mixed-initiative interactive systems as extensions
+built on top of the choose-your-own-adventure calculus. A mixed-initiative interactive system
+defines an operation \ident{suggest} that recommends a sequence of choices for a given state.
+
+\begin{definition}[Mixed-initiative system]
+A choose-your-own-adventure system supports mixed-initiative mode of interaction if it is
+equipped with an operation $\suggest$ such that:
+\begin{itemize}
+  \item $\suggest(\sigma_0)=\iota_1, \ldots, \iota_n$ such that
+    $\forall i\!\in\! 1\ldots n\;.\;\iota_i\mapsto\sigma_i\in\choices(\sigma_{i-1})$.
+\end{itemize}
+\end{definition}
+
+The definition only requires that the suggested sequence of choices is valid, but it
+does not specify how exactly the system should make the recommendations. This is specific to
+a particular system. An interactive theorem prover will try to find a proof or solve
+sub-goals, whereas data wrangling system may try to improve the structure of data.
+In practice, the suggestion should improve the quality of the program so that
+$\select(\sigma_n)$ is better than $\select(\sigma_0)$, but we leave the definition flexible
+to accommodate a broader range of uses.
+
+\subsection{Language model-based completion}
+Large language models can assist with data exploration by generating snippets of code based
+on natural language description of the problem \cite{yin-2023-codegen}. A recognized drawback of
+this approach is that the user may gain only cursory understanding of the code and overlook
+errors. Various systems address this by generating explanations alongside with code \cite{nooralahzadeh-2024-data}.
+We developed a system that provides LLM-based natural language assistance for The Gamma [report
+citation omitted] based on a mechanism that assists the user and is also applicable to other
+choose-your-own-adventure systems.
+
 \subparagraph{LLM assistant for The Gamma.}
 Our system [reprot citation omitted] integrates a type provider like the one used in The Gamma
-with a large language model (GPT-3.5 and GPT-4). It lets user ask a natural language query and
-then recommends choices that the user should make in order to answer the query. Our system does
-not use LLM to generate code, but to recommend an option offered by the type provider. To do so,
-we construct a prompt as shown in Figure~\ref{fig:llm}, which asks the LLM to recommend the next
+with a large language model (GPT, Gemini, DeepSeek). It lets user ask a natural language query and
+then recommends choices that the user should make in order to answer the query. We do not
+use the LLM to generate code, but to recommend an option offered by the type provider.
+
+We construct a prompt as shown in Figure~\ref{fig:llm}, which asks the LLM to recommend the next
 choice. In this example, the LLM is able to reliably respond with ``23'', which is the correct
 choice. Our system then pre-selects it in the completion list offered to the user.
-
 In contrast to one-shot generation of Python or SQL code snippets, our approach guides the user
@@
 
-In addition to the basic prompting strategy outlined above, we evaluated a strategy where the LLM
+In addition to the basic prompting strategy, we evaluated a strategy where the LLM
 is provided with lookahead information, i.e., for each of the choices, we also include a list of
 the choices that will available subsequently (formatted as either inline information in parentheses
-or as a nested bullet-point list). We use two different data sources to evaluate the system,
-each with 10 queries for which we manually determine the correct choices. To compare the results,
+or as a nested bullet-point list). We evaluated the sytem using queries about the Eurostat
+database with 10 queries for which we manually determine the correct choices. To compare the results,
 we compute a score for each query by following the correct path, asking for an LLM recommendation
 in each step and computing the ratio of correct choices. Figure~\ref{fig:methods} shows the
-average scores for five different strategies, using two different LLM engines.
+average scores for five different strategies, using four different LLM engines.
 
 \begin{figure}[t]
-\includegraphics[width=0.7\textwidth]{fig/llm.png}
-\caption{A plot showing the average scores for two different large language models (GPT-3.5 and GPT-4)
-solving 20 challenge problems using five different prompting strategies.}
+\vspace{-1em}
+\includegraphics[width=0.62\textwidth]{fig/llm.png}
+\caption{A plot showing the average scores for four different large language models
+solving 10 challenge problems using five different prompting strategies.}
 \label{fig:methods}
@@
 The results suggest that, when using a more advanced LLM system, the ratio of correct
-recommendations is high-enough to be practically useful. It also shows that providing more
-information to the LLM through the lookahead mechanism improves the quality of the recommendation,
-in a consistent way across multiple LLM systems. Arguably, offering a recommendation is preferrable
-over providing an opaque solution, because it keeps the human in the loop \cite{sellen-2024-copilot}
-and avoids ``ironies of automation'' \cite{bainbridge-1983-ironies} including deskilling.
+recommendations is high-enough to be practically useful. Interestingly,  providing more
+information to the LLM through the lookahead mechanism improves the quality for GPT-based systems,
+but not for other evaluated LLMs. Arguably, the mode of interaction where the user is offered
+a recommendation is preferrable over providing an opaque solution, because it keeps the human
+in the loop \cite{sellen-2024-copilot} and avoids ``ironies of automation''
+\cite{bainbridge-1983-ironies} including deskilling.
 
@@
 
-It is interesting to note that the choose-your-own-adventure calculus provides a common structure
-that makes it possible to integrate multiple different approaches to assisting human in program
-construction, ranging from automatic search (if we can provide a suitable heuristic) to
-the use of LLMs (if generated choices carry enough semantic information). It can also serve as the
-basis for visual programming environments discussed next.
+It is interesting to note that the system sketched here combines a symbolic component (used for generating
+possible choices and checking their correctness) with an AI-based component (used for
+making choices). This is the same architecture that has been used by the Alpha\-Geometry
+system \cite{trinh-2024-alphageometry} for solving geometry problems.
 
@@
 by interacting with the live previews. For this to be possible, the live preview needs to
-offer links to all the possible options offered by $\choices$. This requires a careful design
+offer links to all the options offered by $\choices$. This requires a careful design
 of live previews. In Histogram (Figure~\ref{fig:histogram}), the live preview offers interactive
@@
 key (Figure~\ref{fig:groupby}), the preview shows the key column, but other columns are hidden.
-Specifying an aggregation thus requires an additional (somewhat arbitrary) ``+'' button that
-provides access to the available choices.
+Specifying an aggregation requires an additional ``+'' button that provides access to further choices.
 
 \begin{figure}[t]
-\includegraphics[width=0.6\textwidth]{fig/thegamma3.png}
+\vspace{-1em}
+\includegraphics[width=0.575\textwidth]{fig/thegamma3.png}
 \caption{Specifying aggregation in The Gamma through a live preview. Grouping keys can be
@@
 
-TODO: Is this too general to be practically useful? E.g. LLM for interactive theorem provers
-arguably needs more specific information.
-
-\noindent
-TODO: Something about unnecessary seqentialization
-
-~
-
-\noindent
-TODO: What if the list of choices is infinite?
-
-~
-
-\noindent
-TODO: How do you go back and revisit earlier choices? Backtracking in theorem provers or
-fixing earlier part of the query in data exploration. Sometimes, you can retrace the
-steps, but they may change...
-
-~
-
-\noindent
-Related work - mbark for emacs, VS code command palette
-
-~
-
-\noindent
-Alphageometry paper - follows a similar way of working to choose-your-own-adventure
+Arguably, the study of interactive programming systems is less well-developed than the study
+of programming languages. Our work is a step towards remedying of the situation, but it is
+only one step. As all small formal models, our system ignores a number of practical concerns.
+
+\subparagraph{Revisiting earlier choices.}
+Programmers often revise programs they wrote previously. For example, a data
+scientist may want to add further aggregation to a grouping or change the sorting key.
+The choose-your-own-adventure calculus does not model how such modifications are done.
+Modifying an earlier choice thus poses an interesting problem -- if an earlier choice is
+changed, we may try to ``replay'' the remaining choices based on the identifiers $\iota$,
+but there is no guarantee that they will be offered again and that this will yield the
+desired result.
+
+\subparagraph{Sequentialization of completion.}
+In a number of examples, an expression contained multiple holes or non-terminals that can
+all be completed using the choose-your-own-adventure interaction. We generally assume that they
+can be completed sequentially, but this may be inconvenient limitation. It may be desirable to
+define a variant of expression completion where multiple independent recommendations are
+available for different parts of the program.
+
+\subparagraph{Searching through options.}
+In some systems, the number of options to choose from may be large or infinite. This can be the
+case for complex systems, as well as for correct systems generated from eventually correct systems.
+Similarly, systems may generate a very long chain of small number of options. In those cases,
+choosing options from a menu may be inconvenient -- an alternative interface may lets the
+user search the tree of options.
+
+\subparagraph{Richer user interfaces.}
+The choose-your-own-adventure calculus models a simple interaction based on menus. Although
+we point out that it can be used as the basis for richer interfaces (illustrated by the
+previews in programming by demonstration), it may be interesting to look at richer interfaces.
+In particular, a number of systems are centred around entities that can be selected and
+modified through commands or interactions. This includes Morphic's halos
+\cite{maloney-1995-morphic,maloney-2001-morphic}, the classic Alternate Reality Kit \cite{smith-1986-ark},
+as well as text-based systems \cite{camarena-2025-embark}.
 
 \section{Conclusions}
-
-TODO: Some conclusions
-1) lets us ask all those questions
-2) let ideas cross between domains
+Working in many interactive programming environments has the same feel as following a
+choose-your-own-adventure book. You start and repeatedly choose one of the offered options to
+construct a program, proof or to explore data. The aim of this paper is to formally capture
+this kind of interaction. As with formal models of programming languages, our model focuses on
+the minimal essence of the interaction pattern. Yet, this is sufficient to recognize similarities
+across a broad range of systems, talk about key properties that they may have, transfer knowledge
+across multiple domains, as well as to suggest a more general way of building richer programming
+experiences ranging from AI-assistants and mixed-initiative interaction to programming by demonstration.
 
